---
title: "Very basics of R coding"
author: "Chenxin Li"
date: "9/12/2020"
output: html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Exerice
Today you have learned some basic syntax of R. Now it's time for you to practice. 

Q1: 
1: Insert a new code chunk;
2: make this matrix:
|1|1|2|2|
|2|2|1|2|
|2|3|3|4|
|1|2|3|4| 
and save it as an item called my_mat2
3: select the 1st and 3rd rows and the 1st, 2nd and 4th columns, and save it as an item. 
4: take the square root for each member of my_mat2, then take log2(), and lastly find the maximum value. Use the pipe syntax.  
The command for maximum is max().

```{r}
my_mat2 <- rbind(
  c(1, 1, 2, 2),
  c(2, 2, 1, 2),
  c(2, 3, 3, 4),
  c(1, 2, 3, 4)
)

my_mat3 <- my_mat2[c(1,3), -3]
my_mat3

my_mat2 %>% 
  sqrt() %>% 
  log2() %>% 
  max()
```

Q2: 
1: Insert a code chunk;
2: Use the following info to make a data frame and save it as an item called grades.
Adel got 85 on the exam, Bren got 83 and Cecil got 93. Their letter grades are B, B, and A, respectively. 

3: Pull out the column with the scores. Use the $ syntax. 

```{r}
grades <- data.frame(
  name = c("Adel", "Bren", "Cecil"),
  score = c(85, 83, 93),
  grade = c("B", "B", "A")
) 

grades

grades$score
```










