---
title: "Data Arrangement answer key"
author: "Chenxin Li"
date: "6/19/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction
The first important aspect of data analysis is to get the data in the correct format. 
This is sometimes referred to as getting the data into the tidy format. 
A tidy dataset is a prerequisite for the downstream data visualization and statistical analyses. 
Almost always, the first thing you would do when you start a data analysis is to tidy the data first. 

In this lesson, you will learn how to use the dplyr and tidyr packages to achieve tidy datasets from a "raw" dataset, 
as well as basic operations to data tables.  
We will cover:

1) Basic concepts of tabular data and what is a tidy dataset? 
2) Basic operations to a data table: select, filter, mutate, group and summarise;
3) Basic operations to multiple tables: binding, joining

As a resourse, you can download the data wrangling cheat sheet here for your future references:
https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf 


#load packages
You might need to install some of these packages 
```{r}
library(tidyr) #data arrangement
library(dplyr) #data arrangement 
library(readr) #readling tab, comma and space separated tables into R
library(readxl) #reading excel into R 
library(ggplot2) #R plotting engine 
```


```{r}
child_mortality <- read_csv("data/child_mortality_0_5_year_olds_dying_per_1000_born.csv") 
babies_per_woman <- read_csv("data/children_per_woman_total_fertility.csv") 
```
 
```{r}
head(child_mortality)
head(babies_per_woman)
```
 
```{r}
babies_per_woman_tidy <- babies_per_woman %>% 
  gather("year", "birth", c(2:302)) 

head(babies_per_woman_tidy)

child_mortality_tidy <- child_mortality %>% 
  gather("year", "death_per_1000_born", c(2:302)) 

head(child_mortality_tidy)
```

 
```{r}
birth_and_mortality <- babies_per_woman_tidy %>% 
  inner_join(child_mortality_tidy, by = c("country", "year"))

head(birth_and_mortality)
```
 
 

#Exercise 
You have learned data arrangement! Let's do an exercise to pracise what you have learned today. 
As the example, this time we will use income per person dataset from Gapminder foundation. 
```{r}
setwd("C:/Users/cxli9/Desktop/Li/PBI/PBI299 (Sundar Lab)/Remote_R_learning_v2")
income <- read_csv("data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") 
head(income)
```

##Tidy data
Is this a tidy data frame?
Make it a tidy data frame using this code chunck. 
Hint: the years are spreat out from columns 2 to 242 
```{r}
income_long <- income %>% 
  gather("year", "income", 2:242)

head(income_long)
```

##joining data 
Combine the income data with birth per woman and child mortality data using this code chunck
Name the new data frame birth_and_mortality_and_income
```{r}
birth_and_mortality_and_income <- birth_and_mortality %>% 
  inner_join(income_long, by = c("country", "year"))

head(birth_and_mortality_and_income)
```

##filtering data 
Filter out the data for Bangladesh and Sweden, in years 1945 (when WWII ended) and 2010.
Name the new data frame BS_1945_2010 
How has income, birth per woman and child mortality rate changed during this 55-year period? 
```{r}
BS_1945_2010 <- birth_and_mortality_and_income %>% 
  filter(year == 1945 |
           year == 2010) %>% 
  filter(country == "Bangladesh" |
           country == "Sweden")

BS_1945_2010
```


##mutate data 
Let's say for countries with income between 1000 to 10,000 dollars per year, they are called "fed",
For countries with income above 10,000 dollars per year, they are called "wealthy".
Below 1000, they are called "poor".  

Using this info to make a new column called "status". 
Hint: you will have to use case_when() and the "&" logic somewhere in this chunck. 
```{r}
birth_and_mortality_and_income <- birth_and_mortality_and_income %>% 
  mutate(status = case_when(
    income < 1000 ~ "poor",
    income >= 1000 & income < 10000 ~ "fed",
    income >= 10000 ~ "wealthy"
  ))
```

##summarise the data 
Let's look at the average child mortality and its sd in year 2010 across countries across different status that we just defined 
Name the new data frame child_mortality_summmary_2010
```{r}
child_mortality_summmary_2010 <- birth_and_mortality_and_income %>% 
  filter(year == 2010) %>% 
  group_by(status) %>% 
  summarise(
    mean = mean(birth),
    sd = sd(birth)
  )

child_mortality_summmary_2010
```
 
How does child mortality compare across income group in year 2010? 

Wealthy countries have the least child mortality, followed by fed countries and poor countries have the most. 







