---
title: "Formatting ggplots key"
author: "Chenxin Li"
date: "6/20/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

You might need to install some of these packages

```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
```

# Exercise - Now time for you to try

Let's use the R built-in dataset CO2 as example: 
Ref: Potvin, C., Lechowicz, M. J. and Tardif, S. (1990), Ecology, 71, 1389--1400.

```{r}
head(CO2)
str(CO2)
```

Task: produce a plot that includes the info for Type and Treatment, with uptake on Y axis.
Let's say we are interested in the effect of "Treatment" in each Type.

Hint:

1.  What should be on the x axis?
2.  What variable do you color or fill with?
3.  What faceting layout do you want to use?

Now try to make the best plot you can!


```{r}
CO2 %>% 
  ggplot(aes(x = Treatment, y = uptake)) +
  facet_grid(. ~ Type, switch = "x") + 
  geom_point(aes(fill = Treatment), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  stat_summary(geom = "errorbar", fun.data = mean_se, width = 0.2) +
  scale_fill_manual(values = c("burlywood3", "slateblue3"),
                    limits = c("nonchilled", "chilled")) +
  labs(x = "Type") +
  theme_minimal()+
  theme(legend.position = "top",
        axis.line = element_line(size = 1.2),
        panel.spacing.x = unit(1, "lines"),
        text = element_text(size = 12, color = "black", face = "bold"),
        axis.text.x = element_blank(),
        axis.text = element_text(size = 12, color = "black", face = "bold"),
        strip.placement = "outside"
        )
```
