---
title: "Formatting ggplots key"
author: "Chenxin Li"
date: "6/20/2020"
output:
  html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

The ggplot package in R is a powerful plotting tool. 
It's easy to use, but you can make complex plots with rather simple syntax. 

"gg" stands for graphical grammar - 
It means different layers of objects and elements are layered on top of the previous to generate the plot.
You can find the ggplot cheatsheet here: https://rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf 

This workbook will go over different features of ggplot
1) geoms
2) scales
3) guides
4) themes 
5) facets

...and how to make the clearest, prettiest ggplot possible. 

#load packages
You might need to install some of these packages 
```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
```


 
#Exercise - Now time for you to try

Let's use the R built-in dataset CO2 as example
Ref: Potvin, C., Lechowicz, M. J. and Tardif, S. (1990), Ecology, 71, 1389–1400.
```{r}
head(CO2)
str(CO2)
```

Task: produce a plot that includes the info for Type and Treatment, with uptake on Y axis. 
Let's say we are interested in the effect of "Treament" in each Type. 



Hint:

1) What should be on the x axis?
2) What variable do you color or fill with?
3) What faceting layout do you want to use? 


Now try to make the best plot you can!
```{r}
CO2 %>% 
  ggplot(aes(x = Treatment, y = uptake)) +
  facet_grid(. ~ Type, switch = "x") + 
  geom_point(aes(fill = Treatment), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  stat_summary(geom = "errorbar", fun.data = mean_se, width = 0.2) +
  scale_fill_manual(values = c("burlywood3", "slateblue3"),
                    limits = c("nonchilled", "chilled")) +
  labs(x = "Type") +
  theme_minimal()+
  theme(legend.position = "top") +
  theme(axis.line = element_line(size = 1.2)) + 
  theme(panel.spacing.x = unit(1, "lines")) +
  theme(text = element_text(size = 18, color = "black", face = "bold")) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(strip.placement = "outside")
```



 
