---
title: "One-way ANOVA and the compact letter display key"
author: "Chenxin Li"
date: "6/20/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(emmeans) #three new packages you might need to install them 
library(multcomp)
library(multcompView)

library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

# Data

<https://github.com/allisonhorst/palmerpenguins>\
Ref: Gorman KB, Williams TD, Fraser WR (2014) PLoS ONE 9(3): e90081. <https://doi.org/10.1371/journal.pone.0090081>

```{r}
#remotes::install_github("allisonhorst/palmerpenguins") #download the data from Github.
#you might need to install the "remotes" package
library(palmerpenguins)
```

# Data visualization

```{r}
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")

 penguins %>% 
  ggplot(aes(x = species, y = bill_length_mm)) + 
  geom_point(aes(fill = species),
             position = position_jitter(0.1, seed = 666),
             size = 3,
             shape = 21,
             color = "white") +
  scale_fill_manual(values = wisteria[c(1,6,7)]) +
  labs(x = "species", 
       y = "bill length (mm)") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2),
        text = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        legend.position = "none")
```

# Exercise

Now try it out yourself! Compare the bill depth across the three species and find out which one has the widest bill.

## Visualize the data

```{r}
head(penguins)
```

```{r}
penguins %>% 
  ggplot(aes(x = species, y = bill_depth_mm)) + 
  geom_point(aes(fill = species),
             position = position_jitter(0.1, seed = 666),
             size = 3,
             shape = 21,
             color = "white") +
  scale_fill_manual(values = wisteria[c(1,6,7)]) +
  labs(x = "species", 
       y = "bill length (mm)") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2),
        text = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        legend.position = "none")
```

## Set up the linear model and check assumptions

```{r}
model_depth <- lm(bill_depth_mm ~ species, data = penguins)
plot(model_depth)
```

Prety okay

## Run ANOVA and interpret the ANOVA table

```{r}
anova(model_depth)
```

F = 359, and P value is very small. Reject the null hypothesis of depth being equal across three species.

## Run Tukey tests and interpret the contrast

```{r}
estimate_depth <- emmeans(model_depth, pairwise ~ species)
estimate_depth$contrast
```

## Report the compact letter display and interpret the cld output

```{r}
cld(estimate_depth$emmeans, Letters = letters)
```

Gentoo has the least deep bill.\
The other two species are not statistically different from each other.
