---
title: "One-way ANOVA and the compact letter display key"
author: "Chenxin Li"
date: "6/20/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 

#load packages
```{r}
library(emmeans) #three new packages you might need to install them 
library(multcomp)
library(multcompView)

library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

#data

https://github.com/allisonhorst/palmerpenguins 
Ref: Gorman KB, Williams TD, Fraser WR (2014) PLoS ONE 9(3): e90081. https://doi.org/10.1371/journal.pone.0090081 
```{r}
remotes::install_github("allisonhorst/palmerpenguins") #download the data from Github.
#you might need to install the "remotes" package
library(palmerpenguins)
```

```{r}
head(penguins)
str(penguins)
penguins %>% 
  group_by(species) %>% 
  count()
```


#data visualization
 
```{r}
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")

 penguins %>% 
  ggplot(aes(x = species, y = bill_length_mm)) + 
  geom_point(aes(fill = species),
             position = position_jitter(0.1, seed = 666),
             size = 3,
             shape = 21,
             color = "white") +
  scale_fill_manual(values = wisteria[c(1,6,7)]) +
  labs(x = "species", 
       y = "bill length (mm)") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2),
        text = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        legend.position = "none")
```

 
 
```{r}
length_summary <- penguins %>% 
  filter(is.na(bill_length_mm) == F) %>% 
  group_by(species) %>% 
  summarise(
    mean = mean(bill_length_mm),
    var = var(bill_length_mm),
    sd = sd(bill_length_mm),
    n = n()
  )

head(length_summary)
```

#stat
##setting up a liner model 
To perform an ANOVA in R, a good way to start is to set up a linear model first. 
In a one-way ANOVA, it follows the form lm(y ~ x),
where lm stands for linear model, 
y is dependent variable
x is independent variable 

```{r}
model_bill.length <- lm(bill_length_mm ~ species, data = penguins) #don't forget to specify which data using data = argument
```


```{r}
plot(model_bill.length, which = 2)
```
 

```{r}
plot(model_bill.length, which = c(1, 3))
```


##how to interpret an ANOVA table
```{r}
anova(model_bill.length)
```

 
##how to do Tukey test


```{r}
estimate_bill <- emmeans(model_bill.length, pairwise ~ species)
```

 
```{r}
estimate_bill$contrasts
```
 


 
##How do I report the resuls of a Tukey test?
 
```{r}
cld(estimate_bill$emmeans, Letters = letters)
```

 


##publication quality reporting
 
```{r}
#I'm going to use my wisteria palette
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")
```


```{r}
penguins %>% 
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(aes(fill = species),            #You can make a box plot too!
               alpha = 0.8, width = 0.7) +            
  annotate(geom = "text",     #you can use the annotate funtion to add the letter grouping of cld
           label = c("a", "c", "b"),   
           x = c(1, 2, 3),             #you can also do this in PPT or illustrator or Inkscape 
           y = c(48, 60, 57),   
           size = 5, fontface = "bold") +
  scale_fill_manual(values = wisteria[c(1, 3, 6)]) +
  labs(x = "Species",
       y = "Bill length") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```
   


#Exercise 
Now try it out yourself!
Compare the bill depth across the three species and find out which one has the widest bill. 

##visualize the data
```{r}
head(penguins)
```

```{r}
penguins %>% 
  ggplot(aes(x = species, y = bill_depth_mm)) + 
  geom_point(aes(fill = species),
             position = position_jitter(0.1, seed = 666),
             size = 3,
             shape = 21,
             color = "white") +
  scale_fill_manual(values = wisteria[c(1,6,7)]) +
  labs(x = "species", 
       y = "bill length (mm)") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2),
        text = element_text(size = 18, color = "black", face = "bold"),
        axis.text = element_text(color = "black"),
        legend.position = "none")
```


##set up the linear model and check assumptions
```{r}
model_depth <- lm(bill_depth_mm ~ species, data = penguins)
plot(model_depth)
```
Prety okay 

##run ANOVA and interpret the ANOVA table
```{r}
anova(model_depth)
```
F = 359, and P value is very small. Reject the null hypothesis of depth being equal across three species. 


##run Tukey tests and interpret the contrast
```{r}
estimate_depth <- emmeans(model_depth, pairwise ~ species)
estimate_depth$contrast
```

##report the compact letter display and interpret the cld output
```{r}
cld(estimate_depth$emmeans, Letters = letters)
```
Gentoo has the least deep bill. 
The other two species are not statistically different from each other. 


 






