---
title: "Data transformations key"
author: "Chenxin Li"
date: "6/19/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes   
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

```{r}
C_elegans_lethality <- read_excel("../data/C_elegans_lethality.xlsx")
head(C_elegans_lethality)
```

# Exercise

Now you have learned Wilcoxon tests and data transformation, it's time for you to practice!

## Wilcoxon test

Run pairwise wilcoxon tests on the C elegans lethality data and compare the results to that of the logit transformed Tukey tests.

```{r}
pairwise.wilcox.test(x = C_elegans_lethality$proportion, g = C_elegans_lethality$treatment)
```

Are the your conclusions consistent between the two methods?

Yes, the mean of the treatments are all statistically different from each other.

## Data transformation

Here are the results from an experiment.

```{r}
stem_data <- read_excel("../data/LU-Matand - Stem Data.xlsx")
head(stem_data)
```

In this experiment, the researchers were lookling at the effect of explant in daylily tissue culture.
(Data from Matand et al. 2020. BMC Plant Bio)


They took three types of explant (cut stem, inverted stem and split stem),
put those on agar plates and examined how well shoots regrew from them.
At the end of the experiment,
they counted the number of shoots from each tissue (Buds_Shoots column in the table).
There were other factors in this experiment, but let's ignore them for now.
They did this for a total of 19 daylily cultivars.
For the sake of this exercise, let's just use one of the cultivars - "Alias".

```{r}
Alias <- stem_data %>% 
  filter(Variety == "Alias")  

head(Alias)
```

### Visualize the data using Explant on x axis and Buds_Shoots on y axis.

Label the y axis "Number of regenerated shoots"

```{r}
Alias %>% 
  ggplot(aes(x = Explant, y = Buds_Shoots)) +
  geom_point(aes(fill = Explant), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(y = "Number of regenerated shoots") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )
```

Looking at the plot, do treatments have even spread?
No. Split stem has a lot spread.

### Calculate the summary statistics using group_by() and summarise() for Buds_Shoots

```{r}
 Alias %>% 
  group_by(Explant) %>% 
  summarise(
    mean = mean(Buds_Shoots),
    var = var(Buds_Shoots)
  )
```

Are the variances of treatments within a two-fold range?
No.

### Linear model

Which data transformation would be helpful for this experiment?
Check the assumptions (normality & equal variance) of BOTH original AND transformed data.
Does the data transformation help?

```{r}
#set up linear models in this chunck
model_1 <- lm(Buds_Shoots ~ Explant, data = Alias)
model_log <- lm(log10(Buds_Shoots) ~ Explant, data = Alias)
```

```{r}
#check assumptions in this chunch
plot(model_1, which = c(1, 2, 3))
```

```{r}
plot(model_log, which = c(1, 2, 3))
```

The log-transformed model looks better.

### Interpretation

Run ANOVA and Tukey tests for BOTH original AND transformed data.
Did data transformation (or the lack of) change your conclusions?

```{r}
anova(model_1)
anova(model_log)
```

```{r}
estimate_1 <- emmeans(model_1, pairwise ~ Explant)
estimate_log <- emmeans(model_log, pairwise ~ Explant)

estimate_1$contrast
estimate_log$contrast
```

Essentially the same results for no transformation and log transformation.

### Reporting

Report the estimated means and confidence intervals for the treatments in the cld format.
Don't forget to de-transform if you go with a data transformation.

Because data transformation didn't change the conclusion at all,
for reporting I'm going with the un-transformed model

```{r}
cld(estimate_1$emmeans, Letters = letters)  
```
