---
title: "Randomized Block Design key"
author: "Chenxin Li"
date: "6/21/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
#the same ones we used last time
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

# Visualize data

```{r}
head(OrchardSprays)
```

```{r}
OrchardSprays %>% 
  ggplot(aes(x = treatment, y = decrease)) +
  geom_point(aes(fill = treatment), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(y = "decrease in pests") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )
```

# Exercise

Now you have learned how to analyze randomized block experiments.
Let's practice that!

This time we will look at an actual experiment

```{r}
chicken_experiment <- read_csv("../data/betacarotenoid.csv")
head(chicken_experiment)
```

The researcher was testing the effect of different pastures (P1 to P5) on the amount of beta-carotene in egg.
There were 4 fields (F1 - F4). The fields are the blocking factors.
In each field, hens were randomly assigned to be fed with one of the 5 pastures.
At the end of the experiment, eggs were collected and beta-carotene concentration was assayed.
(Data from UC Davis Plant Sciences course PLS205, Winter 2017)

## Visualize the data

make the best plot you can to visualize the data.

What to put on x axis?
What to color/fill with?
You don't have to include blocking factors in the plot

```{r}
chicken_experiment %>% 
  ggplot(aes(x = Pasture, y = Beta)) +
  geom_point(aes(fill = Pasture), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(y = "Betacarotene content") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )
```

## Setting up the model

Fill out the table:

What are the sources of variation?
What factors are they? (treatment? block? observational units? experimental units?)
How many levels in each?

| source |  factor   | levels |
|:------:|:---------:|:------:|
| Pature | treatment |   5    |
| Field  |   block   |   4    |
|  Egg   |  EU, OU   |   20   |

What should you include in the model? Set up the linear model.

```{r}
model_beta <- lm(Beta ~ Pasture + Field, data = chicken_experiment)
```

## Check assumptions of ANOVA

```{r}
plot(model_beta, which = c(1, 2, 3))
```

Normality?
Pretty good.
Equal variance?
There might be a bit more variance at the higher end of fitted values.
You can try a log transformation, but I don't think it is very critical here.

## Run ANOVA

How would you interpret the ANOVA table?

```{r}
anova(model_beta)
```

Fields did not contributed much to the variation.
The means are statistically different across different pastures.

## Run Tukey tests and show cld

```{r}
estimate_beta <- emmeans(model_beta, pairwise ~ Pasture)
cld(estimate_beta$emmean, Letters = letters)
```

Which pasture treatment gave the best beta-carotene content?

P5 is the highest, although not significantly different from P4 and P3, which all shares the letter c.
