---
title: "Multifactorial Design and interactions key"
author: "Chenxin Li"
date: "6/21/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

# Exercise

Now you have learned how to analyze multifactorial experiment.
You will practice it.

We will use an actual experiment "potato yield" as example.
In this experiment, we are looking at the yield of two varieties of potato across different nitrogen input.

Potato from one of the two varieties were randomly assigned one of the 5 nitrogen treatments.
At the end of the experiment, potatos were harvested and weighed.
(Data from UC Davis Plant Sciences course PLS205, Winter 2017)

```{r}
Potato_Yield <- read.csv("../data/Potato_Yield.csv")
head(Potato_Yield)
str(Potato_Yield)
```

## Visualize the data

make the best plot you can to visualize the data.

What to put on x axis?
What to color/fill with?
Any faceting that you want to use?

WARNING: are there any numerically recorded variables you want to change into factors?
(Hint: there is 1 variable that you need to change into factor.)

Make your plot here:

```{r}
Potato_Yield %>% 
  ggplot(aes(x = Nitrogen, y = Yield)) +
  geom_point(aes(fill = Variety), size = 3, color = "white", shape = 21,
             position = position_jitter(3, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(y = "Yield") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black")
  )
```

## Setting up the model

Fill out the table:

What are the sources of variation?
What factors are they? (treatment? block? observational units? experimental units?)
How many levels in each?

|  source  |  factor   | levels |
|:--------:|:---------:|:------:|
| Variety  | Treatment |   2    |
| Nitrogen | Treatment |   5    |
| potato   |  EU, OU   |  100   |

(The potato yield table has 100 rows.
Because it is a tidy table,
I know there is total of 100 potatoes.)

What main effects should you include in the model?
Which interactions should you include in the model?

Set up the linear model.

```{r}
Potato_Yield <- Potato_Yield %>% 
  mutate(Nitrogen.F = as.factor(Nitrogen)) #don't forget to change numeric variable to factor for ANOVA

model_potato <- lm(Yield ~ Variety * Nitrogen.F, data = Potato_Yield)
```

## Check assumptions of ANOVA

```{r}
plot(model_potato, which = c(1, 2, 3))
```

Normality?
Pretty good.
Equal variance?
Pretty good.

## Run ANOVA

How would you interpret the ANOVA table?

```{r}
anova(model_potato)
```

Variety explains the most variation (F = 140),
followed by Nitrogen (as a factor, F= 21),
followed by their interaction (F = 7)

## Tukey test

Say we are interested in knowing which variety performs better each nitrogen input.
How should you set up the Tukey test?

```{r}
estimate_potato_per_nitrogen <- emmeans(model_potato, pairwise ~ Variety | Nitrogen.F)
```

Pull out the cld output and interpret the results.

```{r}
cld(estimate_potato_per_nitrogen$emmeans, Letters = letters)
```

Across all 5 nitrogen treatment, Russet Burbank has higher yield than Alpine Russet.
