---
title: "Repeated measures keys"
author: "Chenxin Li"
date: "6/21/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(lme4)
library(lmerTest)   #two new packages


library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

# Exercise

Now you have learned how to deal with repeated measures in an experiment,
it's time for you to practice.

We will use a mint growth experiment as an example.
In this experiment, the researchers looked at the effect of 6 treatments (A - F) on mint growth.
Each treatment has 3 pots.
Each pot has 4 plants. The treatment was applied to the pot, not the individual plants.
The growth of each plant in each pot was measured.
(Data from UC Davis Plant Sciences course PLS205, Winter 2017)

```{r}
mint_data <- read_csv("../data/Mint.csv")
head(mint_data)
tail(mint_data)
```

There is a problem with how the data were recorded. Maybe it's not obvious to you.
In each treatment, the name of the Pot is all Pot_1, Pot_2 and Pot_3.
However, in the actual experiment, Pot_1 from Treatment A is not the same Pot_1 for Treatment B, or CDEF.
The same applies to Pot_2 and Pot_3.
Simply put, the pots in each treatment need to have their own unique identifier.
Let's fix that first

```{r}
mint_data <- mint_data %>% 
  mutate(Pot.ID = paste(Pot, Trtmt))

head(mint_data)
```

Now each pot has their own unique ID.

## Visualization

make the best plot you can to visualize the data.

What to put on x axis? What to color/fill with? Any faceting that you want to use?

Make your plot here:

```{r}
mint_data %>% 
  ggplot(aes(x = Trtmt, y = Growth)) +
  geom_point(aes(fill = Trtmt), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(y = "Growth",
       x = "Treatment") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )
```

## What is being repeated measured in this experiment?

Can you tell which entity was repeatedly measured?

Fill in the table and write down the model. What are the sources of variation? What factors are they? (treatment? block? observational units? experimental units?) How many levels in each?

| source |  factor   | levels |
|:------:|:---------:|:------:|
| Treat  | treatment |   6    |
|  Pot   |    EU     |   18   |
| Plant  |    OU     |   72   |

What is the experimental unit for treatment? Pot

What is the observational unit? Plant

Set up the linear model.

## Check assumptions of ANOVA

```{r}
model_mint <- lmer(Growth ~ Trtmt + (1|Pot.ID), data = mint_data)
```

Normality?

```{r}
qqnorm(resid(model_mint))
```

Pretty ok.

Equal variance?

```{r}
plot(model_mint)
```

Fine.

## Run ANOVA

How would you interpret the ANOVA table?

```{r}
anova(model_mint)
```

Means are statistically different across treatments (F = 16, P < 0.05)

## Tukey test

Pull out the cld output and interpret the results.

```{r}
estimate_mint <- emmeans(model_mint, pairwise ~ Trtmt)
cld(estimate_mint$emmeans, Letters = letters)
```

Treatment F are numerically highest,
although not significantly different from and treatments D and E (sharing letter c)
