---
title: "Split field and nested design key"
author: "Chenxin Li"
date: "6/21/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
# load packages
```{r}
library(lme4)
library(lmerTest)  
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```


 

# Exercise
Now you have learned how to handle a split field experiment.   
It's time for you to practice.  

We'll use an oat data as an example.   
In this experiment, we are looking at the effect of fertilizer across 4 variety of oats.    
We have 4 blocks.   
Within each block, it is divided into 4 mainplots. Each mainplot is given a fertilizer treatment (1 - 4).  
Within each mainplot, it is sub-divided into 4 subplots. Each subplot is given a variety of oat (1 - 4).   
At the end of the experiment, oat from each subplot is harvest and the yield is recorded. 
(Data from UC Davis Plant Sciences course PLS205, Winter 2017)    

```{r}
oats_data <- read_csv("data/Oats.csv")
head(oats_data)
```

WARNING: is there any numerically recorded variables you need to change to factors?  
(Hint: there are 4 of them.)  

WARNING: does each mainplot has its unique identifier?  

```{r}
oats_data <- oats_data %>% 
  mutate(Block.F = as.factor(Block)) %>% 
  mutate(Fertilizer.F = as.factor(FertilizerA)) %>% 
  mutate(Variety.F = as.factor(VarietyB)) %>% 
  mutate(Mainplot.F = as.factor(MainPlot))
```


## data visualization

Visualize the data before starting the analysis. Plot Yield on y axis.

What to plot on the x axis?   
what to color/fill with?  
Any faceting you want to use?  
Make the best plot you can here:   
```{r}
oats_data %>% 
  ggplot(aes(x = Fertilizer.F, y = Yield)) +
  geom_point(aes(fill = Variety.F), size = 3, color = "white", shape = 21,
             position = position_jitter(0.1, seed = 666)) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  labs(y = "Yield",
       x = "Fertilizer") +
  theme_minimal() +
  theme(
    axis.line = element_line(size = 1.2),
    text = element_text(size = 14, color = "black", face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "right"
  )
```


## setting up the model 
Fill in the design table.

|source        | factor       | levels |     
|:------------:|:------------:|:------:|   
| Fertilizer   | treatment    |  4     |   
| Variety      | treatment    |  4     |   
| Block        | Block        |  4     |  
| Mainplot     | EU (for fert)|  16    |  
| subplot      | EU/OU        |  64    |   

subplot is the EU for variety   


What are the treatments? Fertilizer and variety    
Any blocking factors? Block  
What is the EU for each treatment?   
EU for fertilizer is mainplot   
EU for variety is subplot   

What is the OU? Subplot  

What should the model be?  
```{r}
model_oats <- lmer(Yield ~ Variety.F * Fertilizer.F + Block + (1|Mainplot.F), data = oats_data )
```


## check assumptions of ANOVA
Normality?
```{r}
qqnorm(resid(model_oats))
```
Pretty good

Equal variance? 
```{r}
plot(model_oats)
```
Fine. 

## run ANOVA and interpret the ANOVA table
```{r}
anova(model_oats)
```
Variety perform similarly across all treatments (P > 0.05)  
Fertilizer has some effect on yield.    
Block has the strongest effect (F = 29). Although as the blocking factor this is not what are interested in.   
The interaction between Variety and fertilizer is also significant (P = 0.03)   

## run Tukey test and identify the best performing oat variety in each fertilizer treatment 
```{r}
estimate_oat_per_fert <- emmeans(model_oats, pairwise ~ Variety.F | Fertilizer.F)
cld(estimate_oat_per_fert$emmeans, Letters = letters)
```
Which variety performed the best?   

There is no statistically significant differences in Fert 2,3 and 4.   
In Fert 1, variety 2 has the best yield, although not significantly differnet from Var 3.   

Comment: Looking at the anova table, it should be obvious that the interesting factor here is Fert, not Var.  
So the more informative Tukey test is to test different across different Fert treatments.    








 
 


 
















