---
title: "Correlation and Regression key"
author: "Chenxin Li"
date: "6/22/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
#load package
```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```


 
```{r}
child_mortality <- read_csv("data/child_mortality_0_5_year_olds_dying_per_1000_born.csv") 
babies_per_woman <- read_csv("data/children_per_woman_total_fertility.csv") 
income <- read_csv("data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")  
```
These are two datasets downloaded from the Gapminder foundation (https://www.gapminder.org/data/). 
The Gapminder foundation has datasets on life expectancy, economy, education, population across countries and years. 
The goal is to remind us not only the “gaps” between developed and developing worlds, 
but also the amazing continuous improvements of quality of life through time.

1) child mortality (0 - 5 year olds) dying per 1000 born.
2) median income per person. 
3) births per woman.
These were recorded from year 1800 and projected all the way to 2100.

```{r}
head(child_mortality)
head(income)
head(babies_per_woman)
```
These tables are not in tidy format. The years spread out across 200+ columns. 
Let's make them into the tidy format first, and then we'll join them by matching the country and year columns.  

```{r}
child_mortality_tidy <- child_mortality %>% 
  gather("year", "death_per_1000", c(2:302)) 

income_tidy <- income %>% 
  gather("year", "income", c(2:242)) 

babies_per_woman_tidy <- babies_per_woman %>% 
  gather("year", "birth", c(2:302)) 

mortality_and_income_and_birth <- child_mortality_tidy %>% 
  inner_join(income_tidy, by = c("country", "year")) %>%
  inner_join(babies_per_woman_tidy, by = c("country","year")) %>% 
  mutate(log10_income = log10(income))

head(mortality_and_income_and_birth)
```
(FYI: income has a huge range and not normally distributed. The best way is to do a log transformation on income first.) 
There are a lot of data. For the sake of this discussion, we'll just look at year 1945 (when WWII ended) for now.


 
 
#Curve fitting using linear model
A limitation of the simple linear regression is that a linear relationship is required. 
However, under certain circumstances, you can also fit curves using linear model.

You can fit curves using linear model when the underlying mathematical relationship is given. 

To understand that, let's use the R built-in dataset Puromycin as an example.
In this experiment, we are looking at the rate of an enzyme across different substrate concentrations. 
There are two states: drug treated or non-treated. We will only use the non-treated control data for now.
(Ref: Treloar, M. A. (1974), Effects of Puromycin on Galactosyltransferase in Golgi Membranes, M.Sc. Thesis, U. of Toronto.) 

```{r}
Puromycin_ctrl <- Puromycin %>% 
  filter(state == "untreated")

head(Puromycin_ctrl)
```


1/V = 2.15e-4*(1/S) + 6.972e-3

So Vmax = 1/6.972e-3, and K = 2.15e-4/6.972e-3

```{r}
1/6.972e-3 #Vmax
2.15e-4/6.972e-3 #K
```

So Vmax = 143, and K = 0.03

In addition, we have an R^2 = 0.89, which means 89% of the variation in the data are explain by the model. Pretty good. 

 
#Exercise one
Now you have learned how to perform correlation test and how to perform linear regression
Let's practise that!

We'll practise correlation test first. We'll use the income and child mortality data again.
This time, we'll use the data from year 2015. 
```{r}
data_2015 <- mortality_and_income_and_birth %>% 
  filter(year == 2015)

head(data_2015)
```

Visualize the relation between birth and child mortality across countries in year 2015.
Make your plot here:
```{r}
data_2015 %>% 
  ggplot(aes(x = birth, y = death_per_1000)) +
  geom_point() +
  labs(x = "Birth per woman",
       y = "deaths per 1000 born") +
    theme_minimal() +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) 
```


Is there a correlation between births per woman and child mortality? Do a correlation test.
```{r}
cor.test(data_2015$birth, data_2015$death_per_1000)
```


What does the correlation coefficient tell you?
There is a strong positive correlation between birth per woman and deaths per 1000 child in the year 2015 

Between income vs mortality in 1945 and birth vs mortality in 2015, which has a stronger correlation? 
```{r}
data_1945 <- mortality_and_income_and_birth %>% 
  filter(year == 1945)
```

```{r}
cor.test(data_1945$log10_income, data_1945$death_per_1000)
```

The magnitude (absolute value) of birth vs. death at 2015 is greater than that of income vs. death at 1945.
Thus a stronger correlation for birth vs. death at 2015. 


#Exercise two
Now let's practice linear regression. We'll use the Puromycin data again.
However, we will use the drug treated data this time.
Again, we are looking at the rate of an enzyme across substrate concentrations, 
but this time the enzyme is treated with the drug puromycin. 

```{r}
Puromycin_treat <- Puromycin %>% 
  filter(state == "treated") %>% 
  mutate(one_over_v = 1/rate) %>% 
  mutate(one_over_s = 1/conc)

head(Puromycin_treat)
```


##set up the model and interpret the model
```{r}
model_puro <- lm(one_over_v ~  one_over_s, data = Puromycin_treat)
summary(model_puro)
```

What is the intercept and slope?
The intersept is 0.0051 and slope is 0.000247

1/v = 0.0051 + 0.000247*1/s

What is the R^2 and what does it mean?
R^2 is 0.84, which means 84% of the variances in the data is explained by the model

Calculate K and Vmax from the coefficients. 
```{r}
1/0.0051    #Vmax
0.000247/0.0051 #K
```

How does puromycin affect the K and Vmax of this enzyme? 

The parameters for the ctrl were: Vmax = 143, and K = 0.03
So puromycin increased both Vmax and K. 


##visualize the model under the double reciprocal scale and the original scale
Make your plots here: 

```{r}
fitted_rate_puro <- data.frame(
  "one_over_s" = seq(min(Puromycin_treat$one_over_s), max(Puromycin_treat$one_over_s), by = 0.1)
) %>%
  mutate(one_over_v = 0.000247*(one_over_s) + 0.0051) %>% 
  mutate(conc = 1/one_over_s) %>% 
  mutate(rate = 1/one_over_v)
```

```{r}
Puromycin_treat %>% 
  ggplot(aes(x = one_over_s, y = one_over_v)) +
  geom_point() +
  geom_line(data = fitted_rate_puro, size = 1.2, color = "indianred") +
  labs(x = "1/S",
       y = "1/V") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) 
```

```{r}
Puromycin_treat %>% 
  ggplot(aes(x = conc, y = rate)) +
  geom_point() +
  geom_line(data = fitted_rate_puro, size = 1.2, color = "indianred") +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```




