---
title: "Logistic Regression key"
author: "Chenxin Li"
date: "6/21/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
library(rsq)  # a new package

library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```

```{r}
qPCR <- read_delim("../data/SYBR_Green.txt", delim = "\t", col_types = cols)
head(qPCR)
```

In this table, each row is a PCR cycle. We have 36 rows, so 36 PCR cycles.
Each column is a PCR reaction. I was doing the experiment in a 96-well plate, so the data are spread out in 96 columns.

Let's visualize the data first

```{r}
qPCR %>% 
  gather("well", "fluorescence", 3:98) %>% 
  filter(fluorescence >= 0) %>%   #remove negative fluorescence - that's just noise or artefact 
  ggplot(aes(x = Cycle, y = fluorescence)) +
  geom_line(aes(group = well, color = well), size = 1, alpha = 0.8) +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.line = element_line(size = 1.2),
        panel.grid = element_line(color = "grey60"),
        text = element_text(size = 12, color = "black", face = "bold"),
        axis.text = element_text(size = 12, color = "black", face = "bold")
        ) 
```

The logit and logistic functions are defined as the following:

```{r}
logit <- function(p){log(
  p / (1-p)
)}

logistic <- function(x){
  1/(1 + exp(-x))
}
```

# Exercise

Now you have learned how to run a logistic regression.
It's time to practice.

This time we'll use the data from qPCR well A7.

## Fit a logistic regression model for well A7

1.  What is Ymax for A7?

```{r}
max(qPCR$A7)
```

It would be 55306.132

```{r}
qPCR_new <- qPCR %>% 
  mutate(A7_scaled = A7/max(A7))
```

2.  What is the equation for A7? What the R^2 and what does it mean?

```{r}
model_A7 <- glm(A7_scaled ~ Cycle, data = qPCR_new, family = binomial(link = "logit"))
```

```{r}
summary(model_A7)
rsq(model_A7, type = "sse")
```

The equation will be:
Y = 1/(1 + e^-(0.3098 * Cycle - 3.2721))

R^2 is 0.988.

3.  Visualize your model. Make your plot here:

```{r}
model_A7_fitted <- data.frame(
  Cycle = seq(1, 36, by = 0.1)
) %>%
  mutate(A7 = max(qPCR$A7) /(1 + exp(-(0.3098 * Cycle - 3.2721))))
```

```{r}
qPCR %>% 
  ggplot(aes(x = Cycle, y = A7)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(data = model_A7_fitted, size = 1.2, color = "indianred1") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.line = element_line(size = 1.2),
        panel.grid = element_line(color = "grey60"),
        text = element_text(size = 12, color = "black", face = "bold"),
        axis.text = element_text(size = 12, color = "black", face = "bold")
        ) 
```

4.  Find the inflection point for A7.

```{r}
3.2721/0.3098
```

It would be 10.56.

```{r}
max(qPCR$A7) /(1 + exp(-(0.3098 * 10.56 - 3.2721)))
1/(1 + exp(-(0.3098 * 10.56198 - 3.2721)))
```

At this point, the y value is ~1/2 of Ymax.
