---
title: "Analyses of proportions, contingency tables and enrichment key"
author: "Chenxin Li"
date: "8/4/2020"
output: html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```


#Excercise 
##Chi-squared goodness of fit test
You did two crosses. 
You crossed a Aa mother to a AA father, you got 30 Aa and 60 AA. 
You crossed a AA mother to a Aa father, you got 55 Aa and 50 AA. 
(The expected proportions should be 1:1.)

Is there distorsion of segregation ratio in either of the crosses? 
If so, which parent is the defect associated with? 

```{r}
chisq.test(x = c(30, 60), p = c(0.5, 0.5))
chisq.test(x = c(55, 50), p = c(0.5, 0.5))
```
There is a significant distortion of segregation ratio in the Aa mom x AA dad cross (p = 0.0016)
This indicates the defects are associated with the maternal parent. 

##Contingency tables and enrichment 
In an experiment, you self an Aa plant. You got 95 F1 offspring. 
89 of them germinated, 6 of them didn't.
Among the germinated ones, you have 16 aa, 51 Aa and 22 AA.
Among the non-germinated (dead) ones, you have 5 aa, 0 Aa and 1 AA. 

Set up a contingency table and test if aa genotype is significantly associated with dead offspring. 
I.e. Is aa seeds enriched in dead seeds? 
           | aa    |  Aa    | AA |  total 
Germinated | 16    |  51    | 22 |  89
Dead       |  5    |  0     |  1 |   6
total      | 21    |  51    | 23 |  95

```{r}
fisher.test(rbind(c(16, 51, 22),
                  c(5, 0, 1)))
```
We have p = 0.0007. 
So aa genotype is preferentially associated with dead seeds. 


Test the same hypothesis using the hypergeometric test. 
```{r}
phyper(q = 5,
       m = 21,
       n = 74, #51 + 23
       k = 6,
       lower.tail = F)
```


Does the conclusions for the tests agree with each other? 
Yes. aa seeds are enriched among dead seeds.
 





 





