---
title: "How to design new color palettes in R"
author: "Chenxin Li"
date: "March 23 2020"
output:
  html_document:
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

Color palette is an important aspect of plots. They make the plots prettier and easy to read. 
The quality of the plots is part of the quality of the work. 
If your plots are nice, people take them more seriously. 
A color palette is a pre-defined set of colors to be used in making plots. 

The desired properties of a color palette include:
1) Colorblind friendly;
2) Easily discernable colors - the colors are not too similar to each other; and 
3) Pretty. 

You can design your own color palette in R. This workbook demonstrates how to do that. 

#load packages
You might need to install some of these packages 
```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(RColorBrewer)
library(viridis)
library(svglite)
library(cowplot) 
```

#data

we will simulate some data
```{r}
data <- data.frame(
  treatment = c("1", "2", "3", "4", "5", "6", "7"),
  response = c(10, 20, 30, 40, 50, 60, 70)
)
```

#Example 1 - "wisteria" 

Wisteria is a color palette that I designed, inspired by the beautiful flowers of the wisteria vines. 
https://en.wikipedia.org/wiki/Wisteria   

If you want a beautiful palette, a good place to start is flowers. 

https://images.homedepot-static.com/productImages/28c9482e-40e2-4b71-b2a3-f43f31409cf5/svn/vigoro-vines-climbing-plants-58023fl-64_1000.jpg



```{r}
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")
```

Most of the experiments have less than 7 treatments, so 7 colors should be good enough. 

R has a bunch of built-in color names. You can directly call their names (in " ") in ggplot, and ggplot will show that color. 
Link to R color names: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf 

You can save the palette as a character string of different color names 


```{r}
data %>%  
  cbind(wisteria) %>% 
  ggplot(aes(x = treatment, y = response)) +
  geom_bar(stat = "identity", aes(fill = treatment), alpha = 0.8) +  #this is alpha = 0.8, or 20% transparent
  geom_text(aes(label = wisteria, y = response), size = 5, hjust = 1.25) +
  scale_fill_manual(values = wisteria) + #call the palette here
  labs(x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 18, face = "bold", color = "black")) +
  theme(axis.text = element_blank()) +
  theme(panel.grid = element_blank()) +
  coord_flip() 
```


```{r}
data %>%  
  cbind(wisteria) %>% 
  ggplot(aes(x = treatment, y = response)) +
  geom_bar(stat = "identity", aes(fill = treatment), alpha = 1) +  #this is alpha = 1, or 0% transparent
  geom_text(aes(label = wisteria, y = response), size = 5, hjust = 1.25) +
  scale_fill_manual(values = wisteria) + #call the palette here
  labs(x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 18, face = "bold", color = "black")) +
  theme(axis.text = element_blank()) +
  theme(panel.grid = element_blank()) +
  coord_flip() 
```

Can you tell the differences between colors? Yes? Good. 

##How do I check if my palette is colorblind friendly? 
There is free software called "Color Oracle" http://colororacle.org/ 
It simulate colorblindness on your computer. 
If you can still discern the colors after you turn on "Deuteranopia", it should be good.  


#Example2 - Higanbana 

Higanbana is another palette I designed, based on the red spider lily. 
https://en.wikipedia.org/wiki/Lycoris_radiata 

https://upload.wikimedia.org/wikipedia/commons/4/48/Red_Spider_Lily----Lycoris_radiata.jpg 

Again if you want beautiful palettes, flowers is a good place to start. 

##pick colors.
What can I do if I don't want to be confined by preexisting colors in R? 

For Windows: "takecolor" is a free app: https://download.cnet.com/TakeColor/3000-2192_4-10387506.html 
After unzipping, run the .exe file. Hover your mouse on the color. Alt + C will record the color. 

For MAC (I don't have a MAC. I hope this works): 
OS X comes with DigitalColor Meter:
Applications > Utilities > DigitalColor Meter.app
It has many options and preferences.
command+shift+c will copy the color under the cursor to the clipboard in many different formats.

```{r}
Higanbana <- c(
  "#FED731",
  "#FDA888",
  "#EF5960",
  "#E70F25",
  "#8C030B",
  "#450110",
  "#1F0A0A" 
)
```

These are the colors I picked from higanbana photos. 
Note: these color names are in the "hex code" format. 
The first two digits are red; second two are green; last two are blue. 
Each digist goes from 0 to F (16-nary), or 0 to 255 (16^2 - 1).  

White is #FFFFFF,
Black is #000000

Red is #FF0000,
Green is #00FF00,
Blue is #0000FF. 

```{r}
data %>% 
  cbind(Higanbana)%>% 
  ggplot(aes(x = treatment, y = response)) +
  geom_bar(stat = "identity", aes(fill = treatment), alpha = 1) + #this is alpha = 1, or 0% transparent 
  geom_text(aes(label = Higanbana, y = response), size = 5, hjust = 1.1) +
  scale_fill_manual(values = Higanbana) + #call the palette here
  labs(x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 18, face = "bold", color = "black")) +
  theme(axis.text = element_blank()) +
  theme(panel.grid = element_blank()) +
  coord_flip()
```

```{r}
data %>% 
  cbind(Higanbana)%>% 
  ggplot(aes(x = treatment, y = response)) +
  geom_bar(stat = "identity", aes(fill = treatment), alpha = 0.8) + #this is alpha = 0.8, or 20% transparent 
  geom_text(aes(label = Higanbana, y = response), size = 5, hjust = 1.1) +
  scale_fill_manual(values = Higanbana) + #call the palette here
  labs(x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(text = element_text(size = 18, face = "bold", color = "black")) +
  theme(axis.text = element_blank()) +
  theme(panel.grid = element_blank()) +
  coord_flip()
```

Can you tell the differences between colors? Yes? Good. 

Again check with Color Oracle to see if your colors are colorblind friendly. 
If you can still discern the colors after you turn on "Deuteranopia", it should be good. 

#How can I design continous color gradients? 

##The viridis package

The best continuous color gradients are always the viridis color maps. 
https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html 

The viridis package contain 4 color scales: viridis, magma, plasma and inferno. 
They are all designed specifically for continuous color gradients.
They are all colorblind friendly, as well as gray-scale friendly, meaning if you print them out in black and white, you can still read them. 

Here's a demostration of the "inferno" option in viridis.

```{r}
library(maps) 
#you might need to install the packages "maps" and "mapproj"
```

```{r}
#load some data. Don't worry about this chunck
unemp <- read.csv("http://datasets.flowingdata.com/unemployment09.csv",
                  header = FALSE, stringsAsFactors = FALSE)
names(unemp) <- c("id", "state_fips", "county_fips", "name", "year",
                  "?", "?", "?", "rate")
unemp$county <- tolower(gsub(" County, [A-Z]{2}", "", unemp$name))
unemp$county <- gsub("^(.*) parish, ..$","\\1", unemp$county)
unemp$state <- gsub("^.*([A-Z]{2}).*$", "\\1", unemp$name)

county_df <- map_data("county", projection = "albers", parameters = c(39, 45)) 

names(county_df) <- c("long", "lat", "group", "order", "state_name", "county")
county_df$state <- state.abb[match(county_df$state_name, tolower(state.name))]
county_df$state_name <- NULL

state_df <- map_data("state", projection = "albers", parameters = c(39, 45))

choropleth <- merge(county_df, unemp, by = c("state", "county"))
choropleth <- choropleth[order(choropleth$order), ]
```

```{r}
ggplot(choropleth, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = rate), colour = alpha("white", 1 / 2), size = 0.2) +
  geom_polygon(data = state_df, colour = "white", fill = NA) +
  coord_fixed() +
  theme_minimal() +
  ggtitle("US unemployment rate by county") +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank()) +
  scale_fill_viridis(option="inferno")
```


##But I still want to make my own! 

Okay, you can try...
```{r}
Higanbana_c <- c(
  "#FFE0A0",
  "#FDA888",
  "#EF5960",
  "#E70F25",
  "#8C030B",
  "#450110",
  "#1F0A0A" ,
  "#000000"   
) %>% 
  rev()
```

It's a good idea to include black (#000000) as the first color, and make the last color lighter. 

```{r}
ggplot(choropleth, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = rate), colour = alpha("white", 1 / 2), size = 0.2) +
  geom_polygon(data = state_df, colour = "white", fill = NA) +
  coord_fixed() +
  theme_minimal() +
  ggtitle("US unemployment rate by county") +
  theme(axis.line = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank(), axis.title = element_blank()) +
  scale_fill_gradientn(colours = Higanbana_c)  #call the colors here
``` 

Well, not too different from viridis-inferno. But I do like the red and black colors. 

#Now design your own palette and try it out! 
Design the best palette you can and send them to Li. 





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
