---
title: "Split field and nested design"
author: "Chenxin Li"
date: "3/26/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction 

Last time we talked about repeated measures, 
and how repeated measures violate the indenpendence assumption of ANOVA.
Hopefully you also realized that, in some experiments, the experimental unit is not the observational unit. 
When that is the case, the experimental unit must be declared as a random effect. 

In this unit, we are going to cover split field/nested design. 
It is an elaboration of randomized block design, but the blocks themselves also serve as experimental units. 
This category probably contains some of the most complex experimental designs you would encounter. 

We will cover:
1) What is a split field experiment/nested design?
2) Why would one set up a split field experiment? 
3) How to set up a model for a split field experiment? 
4) How to analyze a split field experiment? 


#load packages
```{r}
library(lme4)
library(lmerTest)  
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(RColorBrewer)
library(viridis)
library(svglite)
library(cowplot)
```


#What is a split field experiment? 

Consider this experiment: 

The researchers were testing the effect of two pesticide sprays (S1 and S2) on two varieties of rice (V1 and V2).
They had a rice field of 64 spots - 8 rows and 8 columns.  
Each spot had a rice plant.

They used the east side of the field (columns 1-4) to spray S1, and the west side (columns 5-8) to spray S2.
Within each subfield, half of which (odd rows) grew V1, the other half (even rows) grew V2. 
At the end of the experiment, each spot was examined for the decrease of pests.  

The reason they set it up like this is because the spray application covered a large area.
It is impossible to indenpendly apply it to each spot. 
And they used a tracker to shoot the seeds into the field. 
The tracker drove along the row, only switching seeds after a row. 

So how does it look like?

| col 1-4   || col 5-8   |
| spray 1   || spray 2   |  

|V1|V1|V1|V1||V1|V1|V1|V1|   row1
|V2|V2|V2|V2||V2|V2|V2|V2|   row2
|V1|V1|V1|V1||V1|V1|V1|V1|   row3
|V2|V2|V2|V2||V2|V2|V2|V2|   row4
|V1|V1|V1|V1||V1|V1|V1|V1|   row5
|V2|V2|V2|V2||V2|V2|V2|V2|   row6
|V1|V1|V1|V1||V1|V1|V1|V1|   row7
|V2|V2|V2|V2||V2|V2|V2|V2|   row8


This IS a split field experiment. 
The experimental unit for spray is half-field.
However, within each half-field, the EU is further sub-divided into rows, 
which become the EU for variety. 

The characteristic of split field experiment is that 
the EU of one factor is nested within the EU of another factor.  

In this case the EU of variety (rows) is nested within the EU of spray (half-field). 
For this reason, sometimes people refer to split field experiment as "nested design". 


#Why split field? 
Sometimes people set up a split field experiment instead of a multifactorial experiment
because that's the only feasible way.
In the spray example, the spray covers a large area, 
so it is impossible to randomize the spray treatment to each spot.


Other times people set up a split field experiment instead of a multifactorial experiment
because it is more efficient. 

Consider this example:
Researchers were testing the effect of salt (high vs. low) on tissue culture of two genotypes of rice (WT vs. mutant).
They had two petri dishes and each dish had 8 spots. 
Tissues from two genotypes were randomized on each dish, 4 of each genotype per dish. 
One dish was poured with medium containing high salt, the other was poured with medium containing low salt. 
At the end of the experiment, the tissues were examined for the growth.  

So the experimental unit for salt is dish. The experimental unit for genotype is spot. 
Spot is nested within dish. So this is a split field experiment. 

We have 4 reps of each genotpe in each salt treatment. 
When we do a split field experiment, we only need to pour one dish of each salt treatment, instead of four. 
THAT is the power of split field experiment. 
It reduces the number of independent applications, which may be appreciated when the application itself is hard to do,
such as spraying a pesticide or pouring mutliple media. 


#How to set up a linear model for split field? 
The principle is actually pretty simple. 
1) Identify all experimental units;
2) Ask: is this EU also a OU?
   Yes - do not include in the model. We never include OU in models, 
   No - include in the model as a random effect. 
   
That's it. As simple as that.

So let's use the rice spray example. 

| sources | factor | levels | 
| spray   | treat  |   2    | 
| variety | treat  |   2    | 
|subfield | EU*    |   2    | 
| row     | EU*    |   8    |  
| spot    | OU     |   64   |   


Spray and variety are the treatments. 
Subfield is the EU for spray, and not a OU. So it has to be included as a random effect.
Row is the EU for variety, and not a OU. So it has to be included as a random effect too. 
Spot is the OU. At the end of the experiment, it was the spots that were examined for pests. 

So the model will be lmer(pest ~ spray * variety + (1|subfield) + (1|row)) 



Let's use the salt and genotype example now. 

| sources | factor | levels |  
| salt    | treat  |    2   |  
| geno    | treat  |    2   |  
| dish    |  EU*   |    2   |
| spot    | EU, OU |    16  | 

Salt and genotype are the treatments. 
Petri dish is the EU for salt, and not a OU. So it has to be included as a random effect. 
Spot is the EU for genotype, but also the OU. So we don't included it in the model.

Thus the model will be lmer(growth ~ salt * genotype + (1|dish)) 


#How to analyze an split field experiment?
we will use an actual experiment as an example. 
In this experiment, researchers were looking at the effect of temperature and phosphate 
on the leaf thickness across 9 genotypes of maize. 

They had 4 growth chambers (C1 - C4). C2 and C4 were assigned as hot, while C1 and C3 were assigned as cool. 
In each chamber, there were 10 trays (T1 to T10). 
T1 through T5 were watered high phosphate water, while T6 through T10 were watered low phosphate water. 

Each tray contains one plant of each genotpe.
At the end of the experiment, a leaf from each plant was measured for thickness.

It already sounds very complicated. So let's visualize the design in ggplot.

```{r}
maize_data <- read_csv("Maize_thickness.csv")
head(maize_data)
```
```{r}
maize_data <- maize_data %>% 
  mutate(Tray = factor(Tray, levels = c(
    "T1", "T2", "T3", "T4", "T5", "T6", "T7", "T8", "T9", "T10"
  )))
 
maize_data %>% 
  ggplot(aes(x = Tray, y = Geno)) +
  facet_grid(. ~ Chamber) +
  geom_tile(aes(color = Phos, fill = Temp), size = 1.2, alpha = 0.8) +
  scale_color_manual(values = c("grey20", "steelblue4"), limits = c("Low", "High")) +
  scale_fill_manual(values = c("indianred1", "grey90"), limits = c("Hot", "Cool")) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  theme(text = element_text(color = "black", size = 12, face = "bold")) +
  theme(axis.text = element_blank()) 
```

Again four chambers, C1 and C3 were cool; C2 and C4 were hot.
In each chamber, there were 10 trays. The first 5 trays received high Phos, the rest received low Phos. 
Lastly, 9 genotypes placed in each tray. 


##data visualization
Before we analyze the results from this experiment, let's visualize it first. 

Plot Geno on x axis, and thickness on y axis.

Make the best plot you can here:











##setting up the model
It helps to fill out the table before we set up the model.

| source | factor | levels |  
| Temp   | treat  |    2   |
| Phos   | treat  |    2   |
| Geno   | treat  |    9   |
| Chamber| EU*    |    4   |
| Tray   | EU*    |   40!  | 
| plant  | EU, OU |   360  |  


Temp, Phos, and Geno should be starightforward. They are the treatments. 
Chamber is the EU for temperature. It's not an OU. So it needs to be a random effect.
Tray is the EU for Phos. It's also not an OU. So it needs to be a random effect too. 
Plant is the EU for Geno. But plant is also the OU. So we don't include it in the model. 


Maybe you have noticed that there should be 40 trays, not 10. Why?
Because the Tray 1 in chamber 1 is not the same Tray 1 in chamber 2, 3 or 4. 
There should be 10 trays per chamber, so 10 * 4 = 40 trays total. 

Each tray does not have its own unique identifier in the table. we need to fix it. 
```{r}
maize_data <- maize_data %>% 
  mutate(Tray.ID = paste(Chamber, Tray))

head(maize_data)
```
Now each Tray has its own unique ID. 

What will the model be?

```{r}
model_maize <- lmer(thickness ~ Temp * Phos * Geno + (1|Chamber) + (1|Tray), data = maize_data)
```
There is an warning message. 
Not sure why that is, but the rest of the analysis should still run. 


##check ANOVA assumptions
Normality?
```{r}
qqnorm(resid(model_maize))
```

Pretty good!

Equal variance?
```{r}
plot(model_maize)
```
Pretty good!
##ANOVA
```{r}
anova(model_maize)
```

Which treatments explained significant variation in this experiment?
Phos and Geno both explained significant variation in this experiment. 
The means are significantly different across treatments of Phos or Geno.

Which interactions explained significant variation in the experiment? 
Temp:Geno and Phos:Geno. 
This means the effects of Temp & Phos depend on the Geno. 



##Tukey test
This time we have three factors. 
Let's say we are interested in comparing across genotypes under each Phos level under each Temp. 
We will need to set up the emmeans as ~ Geno | Phos | Temp.
It stands for compare across Geno, at each Phos level, as each Temp.

```{r}
estimate_maize <- emmeans(model_maize, pairwise ~ Geno | Phos | Temp)
cld(estimate_maize$emmeans, Letter = letters)
```

We have 4 Temp by Phos combinations (2 * 2 = 4). So even in a cld output, it is getting a bit much to look at. 
 
But there a still few things we can see right away.
1) Under low Phos and cool Temp, all genotype have mean thickness that are not significantly difference from each other. 
2) They are many Geno sharing the same letter in each treatment. 





#Exercise
Now you have learned how to handle a split field experiment. 
It's time for you to practice.

We'll use an oat data as an example. 
In this experiment, we are looking at the effect of fertilizer across 4 variety of oats. 
We have 4 blocks. 
Within each block, it is divided into 4 mainplots. Each mainplot is given a fertilizer treatment (1 - 4).
Within each mainplot, it is sub-divided into 4 subplots. Each subplot is given a variety of oat (1 - 4). 
At the end of the experiment, oat from each subplot is harvest and the yield is recorded. 

```{r}
oats_data <- read_csv("Oats.csv")
head(oats_data)
```

WARNING: is there any numerically recorded variables you need to change to factors?
(Hint: there are 4 of them.)

WARNING: does each mainplot has its unique identifier?


##data visualization

Visualize the data before starting the analysis. Plot Yield on y axis.

What to plot on the x axis? 
what to color/fill with?
Any faceting you want to use?
Make the best plot you can here:











##setting up the model
Fill in the design table.

|source | factor | levels |   
| ...   |  ...   |  ...   | 


What are the treatments?
Any blocking factors?
What is the EU for each treatment?
What is the OU? 

What should the model be?


##check assumptions of ANOVA
Normality?

Equal variance? 


##run ANOVA and interpret the ANOVA table


##run Tukey test and identify the best performing oat variety in each fertilizer treatment 











 
 


 
















Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
