---
title: "Logistic Regression"
author: "Chenxin Li"
date: "3/27/2020"
output:
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction
We talked about linear regression and polynomial regression. 
This time we are going to talk about an extention of linear regression - logistic regression
We will cover:

1) When to perform a logistic regression?
2) How to set up a logistic regression?
3) How to interpret a logsitic regression?


#load packages
```{r}
library(rsq)  # a new package

library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(RColorBrewer)
library(viridis)
library(svglite)
library(cowplot)
```

#When to perform a logistic regrssion and why? 
Not everything in nature occurs in a linear manner. 
If you try to use a linear regression to describe non-linear trend, 
you will get faulty interpretations. 

A very common trend in nature is the logistic curve. 
The logistic curve has many applications in population growth, viral infections, and even PCR. 

Let's use an example. 
Here I was doing a qPCR. In qPCR, a dsDNA dye was added to the reaction mix. The dye is called SYBR green.
SYBR green binds dsDNA and fluoresces. As more PCR cycle occurred, more dsDNA was made, and more flurorescence.
Thus you can use the fluorescence to track the concentration of PCR products. 

```{r}
qPCR <- read_delim("SYBR_Green.txt", delim = "\t")
head(qPCR)
```

In this table, each row is a PCR cycle. We have 36 rows, so 36 PCR cycles. 
Each column is a PCR reaction. I was doing the experiment in a 96-well plate, so the data are spreat out in 96 columns.

Let's visualize the data first
```{r}
qPCR %>% 
  gather("well", "fluorescence", 3:98) %>% 
  filter(fluorescence >= 0) %>%   #remove negative fluorescence - that's just noise or artefact 
  ggplot(aes(x = Cycle, y = fluorescence)) +
  geom_line(aes(group = well, color = well), size = 1, alpha = 0.8) +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(panel.grid = element_line(color = "grey60")) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) 
```

This is the raw data of a qPCR experiment. Look at these nice curves. 
These are clearly non-linear, so if you try to fit a straight line, you would be wrong. 
Each sample starts out as an exponential - fluorescenece increases exponentially as cycle number increaes.
However, at some point, the increase of flurescence slows down, and eventually flattens (we call it saturation). 

Samples with higher template concentration shoots up first and reaches saturation first.
Samples with no temple will be flat lines. 

This is a clear example of logistic curve. 
Logistic curves start out as an exponential, but slow down at some point, and eventually stop increasing.  

Logistic curves have the following equation:
Y = Ymax/ (1 + e^-(ax + b)) 

Ymax is the Y value where the curve reaches saturation. 
"a" determines the steepness of the curve, i.e. how fast the curve shoots up.
"b" determines how far away the curve is from the Y axis, i.e. the horizontal location of the curve. 


The logistic curve is used a lot in modeling population growth. 
At the exponential phase, the pupulation increases rapidly.
However, as resources in the habitat get depleted, the rate of growth slows down.
When resources become scarce, the population size stops growing. 
Ecologists call that "the population has reached the carrying capacity of its habitat". 

The same concept can be applied to PCR.
At the exponential phase, the PCR product increases rapidly, doubling every PCR cycle.
However, as substrates (primers and dNTPs) become depleted, the rate of reaction slows down.
As the end, all primers and/or dNTPs are used up, and the PCR products stop increasing.
Molecular biologist call that "the PCR has reached saturation". 

Lastly, any variables in the range 0 - 1 can be better modeled by logistic regression than linear regression. 
Near 0 or 1, the curve will be in its flat regions, and in between 0 and 1, the curve will be in its exponential region. 

#How to set up a logistic regression
As an example, let's fit a curve for the well A6 of my qPCR experiment,
so our formula will be A6 ~ Cycle. 

To extend linear model to non-linear curves, the linear model have to be extended to a generalized linear model.
We will use generalized linear model function - glm(). 
The underlying mathematics of glm is pretty complex, so we'll focus on the applications, not the theory. 

The glm() function has many applications, but to specify a logistic regression, 
we need to specify glm(... , family = binomial(link = "logit")). 

###find Ymax
First we need to find Ymax. 
It turns out logistic regression only works when the response variable is between 0 and 1.
If the response variable is not between 0 and 1, we need to scale it down to 0 and 1 first. 
This can be easily achieved by dividing every Y value by the maximum Y (Ymax). 

```{r}
qPCR_new <- qPCR %>% 
  mutate(A6_scaled = A6/max(A6))
```

###using the glm() function

```{r}
model_A6 <- glm(A6_scaled ~ Cycle, data = qPCR_new, family = binomial(link = "logit"))
```
The above line sets up the regression model. 
Note that "family = binomial" actually referrs to the fact that now the scale is between 0 and 1.
And "0 vs. 1" is called "binomial".
'link = "logit"', referres to the fact that the inverse function of logistic is logit. 

Y = logistic(ax + b) <=> logit(Y) = ax + b. 
Where logit(Y) = log(Y/1-Y),
where Y is the scaled response variable (between 0 and 1). 




#How to interpret a logistic repression
Again we will use the summary() function to pull out the coefficients
```{r}
summary(model_A6)
```

Looks like the "intercept" is -3.79, and the "slope" for cycle is 0.3486. 
This means logit(Y) = 0.3486 * Cycle - 3.7921. 
Or Y = 1/(1 + e^-(0.3486 * Cycle - 3.7921)). 

You should look at the z values. The null hypothesis is z = 0. 
The farther z is away from 0, the less likely the null hypothesis is true. 
The p-value for both parameters are small, so we can conclude that 
finding these observed parameters is beyond chance. 

To find out the goodness-of-fit in a logistic regression, we will need the rsq() function
```{r}
rsq(model_A6)
```
We have a R^2 of 0.99. Which means 99% of the variation in the data is explained by the model. Exellent. 


#Visualize the model
It's always a good idea to visualize the model before we draw conclusions from it.

```{r}
Ymax_A6 <- max(qPCR$A6)

model_A6_fitted <- data.frame(
  Cycle = seq(1, 36, by = 0.1)
) %>%
  mutate(A6 = Ymax_A6 / (1 + exp(-(0.3486 * Cycle - 3.79)))) 
```

```{r}
qPCR %>% 
  ggplot(aes(x = Cycle, y = A6)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(data = model_A6_fitted, size = 1.2, color = "indianred1") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.line = element_line(size = 1.2)) +
  theme(panel.grid = element_line(color = "grey60")) +
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```
That looks great. 


#predict intermediate values
An important number in logistic curve is the inflection point. 
The inflection point defines where the curve goes from increasing rate to decreasing rate. 

when x < inflection point, the curve is in its exponential phase, and the shape is bending upwards (concave).
When x > inflection point, the curve has left its exponential phase and slowing down, 
and the shape is bending downwards (convex). 

This is a very important number in epidemiology. 
In a viral epidemic, the number of infected cases can be modeled by a logistic curve.
Before the infection point, we know we are going to get more cases every day.
After the infection point, we know are going get less cases every day, 
and the total number of cases will soon stop increasing. 
So being able to know where the inflection point is will help guide how to distribute efforts in an epidemic. 


Let's find the inflection point of our model.
The inflection point is defined the second derivative of the function = 0. f''(x) = 0, and solve for x. 

We have Y = 1/(1 + e^-(0.3486 * Cycle - 3.7921)). If you look at our curve, that should happen between 10 and 12.


If you don't remember calculus, that's ok,
because the x value for the inflection point of y = Ymax/(1 + e^-(ax + b)) is as simple as -b/a... 
Don't believe me? Try out the math yourself. 

```{r}
#inflection point x = 
3.7921/0.3496
```
We get 10.85.
And it is between 10 and 12. See, it works. 

If you plug it in, you will find the y value at the inflection point is Ymax/2. 

```{r}
#Y value at inflection point
Ymax_A6 /(1 + exp(-(0.3486 * 10.85 - 3.7921)))
```

```{r}
Ymax_A6 / (Ymax_A6 /(1 + exp(-(0.3486 * 10.85 - 3.7921)))) 
```

That gives us 2! 

Note that this is a very powerful prediction.
In an on going epidemic, we don't know what the maximal number of cases will be.
However, using existing data, we can guess the inflection point.
The predicted total number of cases will be twice the amount of cases at inflection point 


#Exercise
Now you have learned how to run a logistic regression.
It's time to practise. 

This time we'll use the data from qPCR well A7.


##Fit a logistic regression model for well A7
1) What is Ymax for A7? 



2) What is the equation for A7? What the R^2 and what does it mean?



3) Visualize your model. Make your plot here:



4) Find the inflection point for A7. 








Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
