---
title: "08_Plot_assembly"
author: "Chenxin Li"
date: "2023-07-09"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, we will create publication ready composite plots. 
I.e., each composite plot (or figure) is assembled from multiple panels. 
And the panels themselves could also be a composite of multiple plots. 

You will learn:

1. How to arrange and assemble composite plots.
2. How to control relative heights and widths of panels.
3. How to generate hierarchical composite plots.
4. How to control the panel labels. 

As an example, we will be using the "mtcars" dataset from R. 
Don't worry about what the data or these plots mean. 
This is just an example. 

# Pacakge 
We will be using the `patchwork` package. 
Documentation and tutorials can be found [here](https://github.com/thomasp85/patchwork). 
While it is not the only package for plot assembly, it is my go-to. 

```{r}
library(tidyverse)
library(ggbeeswarm)
library(patchwork)

library(RColorBrewer)
```

# Making the panels 
You can ignore this entire section. 
This section generates the parts that we will use to assemble our final figure. 

## A
```{r}
A2 <- mtcars %>% 
  mutate(AM = case_when(
    am == 1 ~ "manual",
    am == 0 ~ "automatic"
  )) %>% 
  ggplot(aes(x = mpg, y = disp)) +
  geom_point(size = 3, alpha = 0.8, color = "white",
             shape = 21, aes(fill = AM)) +
  scale_fill_manual(values = c("tomato1", "grey20")) +
  labs(x = "Miles/(US) gallon",
       y = "Displacement (cu.in.)",
       fill = "Automatic") +
  theme_classic() +
  theme(
    legend.position = c(0.8, 0.8)
  )

A1 <- mtcars %>% 
  mutate(AM = case_when(
    am == 1 ~ "manual",
    am == 0 ~ "automatic"
  )) %>% 
  ggplot(aes(y = mpg, x = "")) +
  geom_quasirandom(size = 3, alpha = 0.8, color = "white",
             shape = 21, aes(fill = AM)) +
  scale_fill_manual(values = c("tomato1", "grey20")) +
  labs(y = "Miles/(US) gallon",
       x = NULL) +
  theme_classic() +
  theme(
    legend.position = "none"
  ) +
  coord_flip()

A3 <- mtcars %>% 
  mutate(AM = case_when(
    am == 1 ~ "manual",
    am == 0 ~ "automatic"
  )) %>% 
  ggplot(aes(y = disp, x = "" )) +
  geom_quasirandom(size = 3, alpha = 0.8, color = "white",
             shape = 21, aes(fill = AM)) +
  scale_fill_manual(values = c("tomato1", "grey20")) +
  labs(y = "Displacement (cu.in.)",
       x = NULL) +
  theme_classic() +
   theme(
    legend.position = "none"
  ) 

# A_scatter
ggsave(plot = A2, filename = "../Results/08_A2.png", height = 5, width = 5)

# A_hist_mpg
ggsave(plot = A1, filename = "../Results/08_A1.png", height = 3, width = 5)

# A_hist_disp
ggsave(plot = A3, filename = "../Results/08_A3.png", height = 3, width = 5)
```

## B
```{r}
Panel_B <- mtcars %>% 
  mutate(VS = case_when(
    vs == 0 ~ "V shaped",
    vs == 1 ~ "Straight"
  )) %>% 
  ggplot(aes(x = VS, y = hp)) +
  geom_quasirandom(size = 3, alpha = 0.8, 
                   aes(color = VS)) +
  scale_color_manual(values = brewer.pal(8, "Accent")[5:6]) +
  labs(x = "Engine",
       y = "Horsepower") +
  theme_classic() +
  theme(legend.position = "none") +
  coord_flip()

# Panel_B
ggsave("../Results/08_B.png", height = 2, width = 5)
```

## C 
```{r}
Panel_C <- mtcars %>% 
  mutate(rank_wt = rank(wt, ties.method = "first")) %>% 
  ggplot(aes(x = "", y = rank_wt)) +
  geom_tile(aes(fill = mpg)) +
  scale_fill_gradientn(colours = brewer.pal(9, "YlGnBu")) +
  labs(x = NULL,
       y = "Rank of Weight",
       fill = "Miles/(US) gallon") +
  theme_classic()

# Panel_C
ggsave("../Results/08_C.png", height = 6, width = 3)
```


# Putting plots together 
Let's say we want to make a figure with three panels: A, B & C. 
And we want a layout like this: 

| Column 1    | Column 2 |
| :---------: | :------: |
| Panel A     | Panel C  |
| Panel B     | Panel C  |

The top left corner is panel A.
The lower left corner is panel B.
The right side is panel C. 

## Panel A
For the sake of this lesson, let's make panel A itself a composite of three sub-panels. 
The 3 sub-panels are called "A1", "A2", and "A3". 

Let's say I want panel A to look like this: 

| Column 1    | Column 2  |
| :---------: | :-------: |
| A1          | Blank     |
| A2          | A3        |

Notice that we only have 3 sub-panels to fill this 2x2 layout. 
In this case the upper right corner is left blank. 
What should we do? 

Let's look at each sub-panel for panel A first. 
I already saved them as an R item in the memory. 
```{r}
A1
A2
A3
```

### The design of the composite 
To create a composite, we use the `wrap_plots()` function from `patchwork`. 
The name is somewhat self-explanatory as for wrapping plots into a larger one. 
`patchwork` has an intuitive way to arrange plots using the `design = ` argument. 
In this case, the design looks like this: 

```{r}
design <- c(
  "A#
   BC"
)
```

The `design` argument is a character vector of uppercase letters (A, B, C, and so on).
within this vector we start new lines to mean different rows in the composite.
In this case, the `design` has two rows, so the composite will have two rows. 
It specified that "part A" will be at the top left, and that "part B" will be at the lower left. 
Finally, "part C" will be at the lower right. 
The "#" sign indicates a blank space. 

To set it up, we simply call `wrap_plot()` and list the sub-panels in order.
```{r}
wrap_plots(A1, A2, A3,
           design = c("A#
                       BC"))

ggsave("../Results/08_A_1.png", height = 6, width = 6)
```
The way `wrap_plots()` matches sub-panel names to parts A, B, and C is by the order. 
In this case, the first sub-panel name (A1) provided is part A, and so on. 

### Controling relative heights and widths of sub-panels. 
`patchwork` allows you control the relative heights and the widths of sub-panels. 
In this example, we have a 2x2 design. 
Let's say I want the 1st row to 30% of the height of the 2nd row.
And I want the 2nd column to be 30% of the width of the 1st. 
This can be specified using the `heights = ` and `widths =` argument within `wrap_plots()`.

```{r}
wrap_plots(A1, A2, A3,
           design = c("A#
                       BC"), 
           heights = c(0.3, 1),
           widths = c(1, 0.3))

ggsave("../Results/08_A_2.png", height = 6, width = 6)
```
For example, vector `c(0.3, 1)` passed on to the `widths` option specifies that the 1st column should be 0.3 of the width of the 2nd column.  
Notice we only 2 rows and columns, so the heights and widths vectors each have 2 numbers. 
However, if the dimension of the design is different, the heights & widths vectors must have matching dimension too. 
For example, if your design have 3 rows and 4 columns, the heights vector must have 3 numbers, and the widths vector must have 4 columns. 

Finally, we can control where we put panel labels. 
In publication ready figures, each panel is labeled by a letter (e.g., A, B, C and so on).
In this example, panel A has 3 sub-panels, how do we control which sub-panel gets labeled? 
We can control this by using `+ labs(tag = "...")` option in `wrap_plots()`. 
Let's say we want the letter "A" to appear at the upper left of panel A. 
The best way to do this is appending `+ labs(tag = "A")` to the first sub-panel, in this case A1.  


```{r}
Panel_A <- wrap_plots(A1 +
                        labs(tag = "A"), 
           A2, A3, 
           design = c("A#
                       BC"),
           heights = c(0.3, 1),
           widths = c(1, 0.3))

Panel_A
ggsave("../Results/08_A.png", height = 6, width = 6)
```
Now we have panel A! 

## Hierarchical assembly 
We made panel A, but we still need to assemble panel A with panels B and C.
Well, we don't have to stop here. 
We can call `wrap_plots()` again, now using "Panel_A" as one of the parts. 

Here are panels B and C:
```{r}
Panel_B
Panel_C
```

Let's say we want panel A and panel B to be in the left column, panel C alone take up the entire right column, what should the design argument be? 
```{r}
c("AC
   BC")
```
In this case, the design vector has two rows. 
And we specified that part C should take up both rows of the 2nd column, as we specified that part C should occupy both the 1st and 2nd rows. 

And let's say we want the 2nd column to the 1/10 of the width of the 1st.
And that the 2nd row to be 30% of the height of the 1st. 
Again, we can control that by calling `widths = ` and `heights = ` within `wrap_plots()`. 

```{r}
 wrap_plots(Panel_A, 
                     Panel_B, 
                     Panel_C, 
                     design = "AC
                               BC",
                     widths = c(1, 0.1),
                     heights = c(1, 0.3))

ggsave("../Results/08_assembled.png", height = 7, width = 8)
```
Now we are really close. 
Notice that only panel A is labeled? 
We want panel B to be labeled with a "B" and panel C to be labeled with a "C". 
We just need to append `+ labs(tag = "B")` and `+ labs(tag = "C")` to the respective panels. 

## Adding panel label 
```{r}
 wrap_plots(Panel_A, 
              Panel_B +
              labs(tag = "B"), 
                     Panel_C +
              labs(tag = "C"), 
                     design = "AC
                               BC",
                     widths = c(1, 0.1),
                     heights = c(1, 0.3))

ggsave("../Results/08_assembled_label.png", height = 7, width = 8)
```
Done! 
Great, we have a publication ready figure! 

# Exercise
Use your own data from your research to generate a composite figure. 
Save it using `ggsave()`. 

