---
title: "Formatting ggplots"
author: "Chenxin Li"
date: "6/20/2020"
output:
  html_document:
    toc: yes  
  html_notebook:   
    number_sections: yes    
    toc: yes  
    toc_float: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Codes are highlighted in ```grey```. 

# Introduction

The ggplot package in R is a powerful plotting tool.   
It's easy to use, but you can make complex plots with rather simple syntax.   

"gg" stands for graphical grammar -  
It means different layers of objects and elements are layered on top of the previous to generate the plot.  
You can find the [ggplot cheatsheet](https://rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf)  

This workbook will go over different features of ggplot  

1) geoms  
2) scales  
3) guides  
4) themes   
5) facets    

...and how to make the clearest, prettiest ggplot possible.    

# load packages
You might need to install some of these packages 
```{r}
library(ggplot2) 
library(tidyr)
library(dplyr)
```

We'll use the R built-in dataset ToothGrowth.   
This dataset is about different doses of vitamin C on tooth growth of guinea pigs   
Ref: Crampton, E. W. (1947). The Journal of Nutrition, 33(5), 491–504. doi: 10.1093/jn/33.5.491.

```{r}
head(ToothGrowth)
str(ToothGrowth)
```
It has 60 rows and 3 columns.   
Two types of supp: OJ and VC   
Looks like len is tooth length and the dependent variable.    
Let's make the most basic plot first   

# coordinate system 
Dose on X, length on Y, color as supp 
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(color = supp)) 
```
I prefer to type out the data frame first, then pipe into ```ggplot()```. It's more readable this way.    
Thus ```ToothGrowth %>% ggplot(...). ```

The line ```ggplot(aes(x = ..., y = ...))``` specifies what are the x and y axis.  
In this case does on x axis, len on y axis.  
Then you add a new layer (in this case points) by using ```+```     
and in a new line you write ```geom_poits()```. "Geom" stands for geommetric object.   
In this case ```geom_points()``` produces points.   
Inside ```geom_point()``` you want to color by supp, so you write ```aes(color = supp)```.    


This is not bad... Now let's use a custom palette called "wisteria". It's just a set of color names.    

# adjust geom color/fill
```{r}
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")
```
This is just a list of color names in R. 

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(color = supp)) +
  scale_color_manual(values = wisteria[c(1, 7)]) #calling the 1st and 7th color 
```
You can customize color using ```scale_color_manual()```,   
in side which, you specfies the names of the colors,   
in this case the color names are pre-specified by a vector called ```wisteria```.   

# change axis and legend labels - labs() 
Now I want to change the axis and legend titles. How do I do that?  

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(color = supp)) +
  scale_color_manual(values = wisteria[c(1, 7)]) +
  labs(x = "dose",              #the labs option allow you to change axis and legend labels. 
       y = "teeth length",
       color = "supplement")
```
The ```labs()``` function allows you to change axis and legend labels.   
Should be self-explanatory.   

You can see the dots are overlapping each other. How do I make them spread out a bit horizontally?    
# adjust geom positions

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(color = supp), 
             position = position_jitter(0.1),   #horizontal spread of 0.1
             alpha = 0.8,             #20% transparent. When dots do overlap, we can tell
             size = 3) +    #make them larger and easier to see 
  scale_color_manual(values = wisteria[c(1, 7)]) +
  labs(x = "dose",              
       y = "teeth length",
       color = "supplement")
```

There are a few options in geom_points. 
The ```position = position_jitter(0.1)``` argument adds random horizontal spread of 0.1 to each dot.  
The ```alpha = 0.8``` makes your dots 20% transparent. (```alpha = 1``` is 0% transparent and default.)  
The ```size``` option allows you to make dots larger or smaller.  

This is great, but I can make it even better by adding border to each dots. How to do that?  
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(fill = supp), #note that change to fill = supp
             position = position_jitter(0.1),   
             alpha = 0.8,             
             size = 3,
             shape = 21, #shape = 21 gives you open circle
             color = "black") +    #black border around each dot
  scale_fill_manual(values = wisteria[c(1, 7)]) + #change to scale_fill here 
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement") #change to fill = here
```
## color vs fill 
Well there are more options in ```geom_point```.
One of them is shape. ```geom_point``` allows different shapes, such as square, cross and etc.  
My favorite is open circle. To use the open circles, you call ```shape = 21```.   
The advantage of open circle is that it has both ```color``` and ```fill``` properties.   
To help you understand the difference between color and fill, let's compare the following two plots.   

```{r}
data.frame(
  x = c("group1", "group2"),
  y = c(10, 20)
) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_bar(stat = "identity", aes(fill = x)) +
  ggtitle("fill by x")
```

```{r}
data.frame(
  x = c("group1", "group2"),
  y = c(10, 20)
) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_bar(stat = "identity", aes(color = x)) +
  ggtitle("color by x")
```
I hope you can see the difference. 
If we use ```fill```, the entire interior is colored, or filled. 
If we use ```color```, only the outline is colored. 

Coming back to using open circles in ```geom_point()```,  
Open circles takes both ```fill``` and ```color```.    
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(fill = supp), #note that change to fill = supp
             position = position_jitter(0.1),   
             alpha = 0.8,             
             size = 3,
             shape = 21, #shape = 21 gives you open circle
             color = "black") +    #black border around each dot
  scale_fill_manual(values = wisteria[c(1, 7)]) + #change to scale_fill here 
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement") #change to fill = here
```
In this case, I use ```color = "black"``` to make black outlines for each dot.   
Then I used ```aes(fill = supp)``` to fill the interior of dots by the supp.  
Now, to specify colors, I have to use ```scale_fill_manual()```,   
instead of ```scale_color_manual()```,  because the variable ```supp``` is now reflected by fill, not color.  

But there is still a problem...  
Note that when you use jitter, the horizontal spread is completely random.      
This means everytime you run the code, the position of the dots are different.     
How do I fix that?    
You can add a ```seed``` argument within ```position_jitter()```, that will make it reproducible   

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),   #added seed = 666 here
                                                    #doesn't matter which seed number you use
                                                    #as long as it's the same number the pattern will be the same
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement")
```
In this case, I used ```seed = 666```.   
But you can use any seed number. 
As long as it's the same number, the plot will look the same everytime you run the code.   


# stat_summary - very powerful visualization tool   

Let's say I want a curve for each supplement, what should I do?    
Your first reaction is ```geom_line```, no?      
Unfortunately, that's not the right answer.    
```geom_line``` will try to connect each dot,  
but what we want is try to connect the *means* of each dose for each supp.   
We'll use ```stat_summary()``` instead   

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  #draw a line
               fun.y = mean, #summarise by averaging, thus mean
               aes(group = supp, color = supp), #one line per supp, color by supp
               size = 1.2) + 
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement")
```
In ```stat_summary()```, I put ```geom = "line"```, which asked it to draw lines.  
Then I put ```fun.y = mean```, which asked it summarize y values using the ```mean()``` function.  
Then I put ```aes(group = supp, color = supp)```, which asked for one line per supp,   
and color those lines by supp.   

well, there's a problem. The lines are not the same color as the dots.    
That's because we specified the interior colors of dots using ```scale_fill_manual()```,    
but we never specified the color of the lines.  
It's easy to fix. We will use ```scale_color_manual()``` to specify colors for the lines too.   

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + #add scale_color_manual here  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") #change the color legend title here 
```
Better!   
Note that since the title and scales of fill and color are the same, ggplot merged them in the legend.    

Now I want to add errorbars to the line, how do I do that?   
We'll use ```stat_summary``` again,   
but this time instead of ```geom = "line"```, we'll do ```geom = "errorbar"```   
Besides ```geom = "errorbar"```, another option is ```geom = "linerange"```   
linerange is errorbar without the horizontal bars

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "errorbar",  #draw errorbars (another option is geom = "linerange")
               fun.data = mean_se, #mean_se specifies the errorbar to be based on standard error (se)
               aes(group = supp), #one errorbar per dose per supp
               width = 0.1) +   #make the bar narrower 
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement")  
```
Now there are error bars.    
Here is another option to show the errorbar using ```stat_summary(geom = "ribbon")```    
I like ribbon better. You can see for yourself. 

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",  #draw ribbons around line
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   #make ribbons more transparent 
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement")
```
I like the ribbon better.    
It looks very nice for high transparency with ```alpha = 0.5```.     
But feel free to use errorbar or linerange if you prefer those    

# adjust legends and axis

Note that in the experiment, there's never a dose = 1.5.   
We can turn off the x axis tick mark of 1.5 by using ```scale_x```        
(You can adjust y axis using ```scale_y``` as well, if you ever need to)   

BTW, if you ever need to log10 transform axis, you can do ```scale_x_log10()``` or ```scale_y_log10()```    
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) + #use scale_x_continuos here because x is continuos
                                              #if you have discrete x, you will need scale_x_discrete
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement")
```
Because x axis value is numeric and continuous, you will use ```scale_x_continuous()```  
And inside, you set the break makers, in this case 0.5, 1 and 2.   
Now the x = 1.5 mark is turned off.     

I realized the color and fill scales are redundant, so we can turn off one of them.   
To do so, we'll use guides.     

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) #turn off color legend using color = F 
```
The ```guides()``` function can do a lot of stuff. 
Here I turned color legend off by putting ```color = F``` in ```guides()```.     

# adjust themes
The default theme is nice, but it's not the best   
You can use the ```theme()``` function to build your own theme  

Let's remove all the theme element first using ```theme_minimal()```  

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  theme_minimal() 
```

Theme minimal is the minimalistic theme. It turns off most theme elements.   
But you can add elements that you want back.    

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  theme_minimal() +
  theme(axis.line = element_line(size = 1.2)) +  #add axis line back
  theme(text = element_text(size = 12, color = "black", face = "bold")) + #make text clearer and darker
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) 
```
The ```theme()``` function takes a lot of options, such as ```axis.line```, ```text``` and all that.  
You can look up the options [here](https://ggplot2.tidyverse.org/reference/theme.html)   
The rule of thumb is if you want to adjust lines, such as axis line, then you use ```element_line()```   
Then you specify color and size in ```element_line()```   
If you want to adjust text, then you use ```element_text()```,   
again in which you specify color and size in ```element_text()```.   
The white backgroung and black lines really put the data forward and emphasize the data    

Now notice the legend is on the right and it is taking up space?     
You can move it into the plot using ```theme(legend.position = )```
 
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  theme_minimal() +
  theme(legend.position = "bottom") + #move legend to bottom
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) 
```
In ```theme()```, you can put ```legend.position = "bottom"```.   
You can also put ```"top"```, ```"right"``` or ```"left"```.     

The more fancy thing is to move legend inside the graph area. How to do that?    
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2)) + #this is a tricky command, and you have to manually adjust it
                                         #c(1, 1) means upper right corner of the graph area,
                                         #c(1, 0) means the lower right corner of the graph area
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```

In ```legend.position = c(x, y)```, you manually adjust where the legend will show up.   
For example ```c(1, 1)``` means upper right corner of the graph area,  
and ```c(1, 0)``` means the lower right corner of the graph area  

Now the legend is in the graph area.    
This is nice because now the legend doesn't take up extra space    

One last adjustment you can do is change the number of rows and column of the legend.   
Now we have two rows, one column. What if I want one row and two columns?    

We'll use guides again. 
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) + # set fill legend to 1 row 2 columns. 
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2)) + 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```
This line ```guides(fill = guide_legend(nrow = N))``` is very useful we you have limited space
and a lot of legend items    
It also works as ```guides(color = guide_legend(nrow = N))``` if you need to arrange color legends. 
Again, think about the difference between fill and color. 

# faceting
faceting is very important when you have a factorial design.   
In this experiment, we have supp as a factor, and dose as another factor.    
So this is a factorial design   
faceting allows you and readers to focus on the true comparison of interest    

Let's say we are interested in comparing the effect of each supp across different doses...     
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  facet_grid(. ~ supp) + #see comments after plot for explanations 
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2)) + 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```
Now it makes two plots.    
The facet syntax works like this:   
```facet_grid(row ~ column)``` 
We are using supp as columns, and only 1 row, so 
```facet_grid(. ~ supp)```. The ```.``` before the ```~``` is just a space holder.     

You can increase the panel spacing too      
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  facet_grid(. ~ supp) +  
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "none") +  #this time the legend is redundant w/ the panel label, so I turn it off. 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines")) #make 1.2 lines of spacing between the subplots (panels)
```
Note that this time the legend is redundant with the panel label,   
so I turn it off using ```legend.position = "none"``` in ```theme()```    
Great!    

But now say I am actually interested in comparing between supp at each dose, what do I do?   
We'll need ```facet_grid(. ~ dose)```, no?       

```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  facet_grid(. ~ dose) +  #change to ~ dose
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "none") +   
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines"))  
```

Now there is a problem. The lines and ribbons are gone.    
That's because in each panel, there is only one dose. There is nothing to connect.   
And the x axis is redundant with the panel.     
So what we really need is to put supp on x axis instead     

```{r}
ToothGrowth %>% 
  ggplot(aes(x = supp, y = len)) +
  facet_grid(. ~ dose) +    
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  #scale_x_continuous(breaks = c(0.5, 1, 2)) +  remove this. this time x axis is discrete
  labs(x = "supplement",  #now x = supplment              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "none") +  #this time the legend is redundant with the x label, so I turn it off. 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines"))  
```
There is no numeric trend on x axis, so I removed the ribbon and line.   
But I can still add the errorbars back.  

```{r}
ToothGrowth %>% 
  ggplot(aes(x = supp, y = len)) +
  facet_grid(. ~ dose) +    
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") + 
  stat_summary(geom = "errorbar", #draw errorbars
               width = 0.2, size = 1,
               aes(group = supp), #one bar per supp per dose
               fun.data = mean_se) +   #errorbar base on standard error (SE)
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  labs(x = "supplement",  #now x = supplment              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "none") +  
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines"))  
```
Pretty good.   
There is one problem. "oJ" and "VC" text appear multiple times.    
We can turn them off and only show the legend.   

```{r}
ToothGrowth %>% 
  ggplot(aes(x = supp, y = len)) +
  facet_grid(. ~ dose) +    
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") + 
  stat_summary(geom = "errorbar",  
               width = 0.2, size = 1,
               aes(group = supp),  
               fun.data = mean_se) +   
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_discrete(labels = NULL) + #turn off x axis text using labels = NULL 
  labs(x = "supplement",               
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "bottom") +  #put the legends back 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines"))
```

Now just a couple final touches. Note that the text "supplement" appears twice, and dose appears none.    
We can put the panel labels to the bottom, and swith the x axis label to dose instead.   

```{r}
ToothGrowth %>% 
  ggplot(aes(x = supp, y = len)) +
  facet_grid(. ~ dose, switch = "x") + #switch = "x" puts x panel labels to the bottom instead of top    
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") + 
  stat_summary(geom = "errorbar", 
               width = 0.2, size = 1,
               aes(group = supp),  
               fun.data = mean_se) +     
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_discrete(labels = NULL) +  
  labs(x = "dose",  #now x is dose again               
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) +  
  theme_minimal() +
  theme(legend.position = "top") +  #legends to the top, make space for x axis label
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold")) +
  theme(panel.spacing = unit(1.2, "lines")) +
  theme(strip.placement = "outside") # put the text 0.5, 1, and 2  outside the axis line
```
Now it really looks like dose is x axis, but in fact it is the faceting factor.    

Congratulation! You have produced a publication quality plot!    

I think this is the best plot we can make for faceting by dose.    
The best practice of plotting is always idenity the comparison of interest,   
remove redundant and unnecessary elements,     
try to make it as pretty as you can, and      
be clear on what you want to show.      

# Exercise - Now time for you to try

Let's use the R built-in dataset CO2 as example   
Ref: Potvin, C., Lechowicz, M. J. and Tardif, S. (1990), Ecology, 71, 1389–1400.   
```{r}
head(CO2)
str(CO2)
```

Task: produce a plot that includes the info for Type and Treatment, with uptake on Y axis.   
Let's say we are interested in the effect of "Treament" in each Type.    



Hint:

1) What should be on the x axis?
2) What variable do you color or fill with?
3) What faceting layout do you want to use? 

Use the format:   

CO2 %>%    
  ggplot(aes(x = ???, y = uptake)) +   
  facet_grid(??? ~ ???) +    
  geom_ponit(aes()) +   
  stat_summary(geom = errorbar) +   
  scale_color_manual() +   
  scale_fill_manual() +   
  scale_..._.... +   
  labs() +   
  guides() +   
  theme_minimal()+   
  theme(legend.position = ...) +   
  theme(axis.line = element_line(size = 1.2)) +     
  theme(text = element_text(size = 12, color = "black", face = "bold")) +   
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))      

Now try to make the best plot you can!      
```{r}

```



 
