---
title: "Multifactorial Design and interactions"
author: "Chenxin Li"
date: "6/21/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 
#Introduction

Last time we talked about randomized block design, 
and how one can using blocks to control for unintended variations in an experiment. 

This time we are going to talk about multifactorial design. We will cover:

1) What is a multifactorial design? 
2) Why would one design a multifactorial experiment?
3) What is interaction between factors?
4) How to analyze a multifactorial experiment? 

#load pacakges
```{r}
library(emmeans)  
library(multcomp)
library(multcompView)
library(ggplot2) 
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(RColorBrewer)
library(viridis)
```



#What is a multifactorial design? 

A multifactorial design is an experiment that have multiple factors of independent variables. 
It might seem a bit abstract. So let's see an actual example.

We will use the R built-in dataset ToothGrowth as an example.
ToothGrowth is an experiment in which researchers looked at the effect of two supplements "VC" and "OJ" 
on the tooth length of guinea pigs.

Each animal was given a supplement (either OJ or VC);
and the animal can receive a high dose (2), mid dose (1) or low dose (0.5).
At the end of the experiment, the longest tooth of the animal is measured. 
Ref: Crampton, E. W. (1947). The Journal of Nutrition, 33(5), 491–504. doi: 10.1093/jn/33.5.491.

```{r}
head(ToothGrowth)
str(ToothGrowth)
```
Looks like it has 60 rows and 3 columns. 
"len" is tooth length. It will be our dependent variable. 
There are 2 kinds of "supp": "VC" or "OJ"
And there are multiple levels of dose, which is recorded as a numeric variable in the table. 

This experiment is called a multifactorial design because 
it has two independent variables of interest: supp and dose. 

supp is a factor; dose is another factor. We have two factors of interest, thus multifactorial. 

#Why would one want to do a multifactorial experiment? 
The multifactorial design is popular because it is efficient. 
When we have multiple independent variables of interest, 
it allows us to compare multiple combinations of independent variables in the same time. 

Let's use the ToothGrowth data as an example. 
There are two supplements and three doses of each. 
2 * 3 = 6 combinations. 
And the multifactorial design will allow us to compare all 6 combinations in the same time. 


#What is interaction among factors?
A very important concept of the multifactorial design is interaction. 
Interaction stands for "the effect of one factor depends on the other factor".

To understand this, let's visualize the data first. 

```{r}
wisteria <- c("grey65", "burlywood3", "khaki2", "plum1", "lightcyan2", "cornflowerblue", "slateblue3")

ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.y = mean,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "ribbon",   
               fun.data = mean_se, 
               aes(group = supp,
                   fill = supp), 
               alpha = 0.5) +   
  geom_point(aes(fill = supp), 
             position = position_jitter(0.1, seed = 666),    
             alpha = 0.8,             
             size = 3,
             shape = 21, 
             color = "black") +    
  scale_fill_manual(values = wisteria[c(1, 7)]) +  
  scale_color_manual(values = wisteria[c(1, 7)]) + 
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       fill = "supplement",
       color = "supplement") +
  guides(color = F) +
  guides(fill = guide_legend(nrow = 1, ncol = 2)) + # set fill legend to 1 row 2 columns. 
  theme_minimal() +
  theme(legend.position = c(0.8, 0.2)) + 
  theme(axis.line = element_line(size = 1.2)) + 
  theme(text = element_text(size = 12, color = "black", face = "bold")) +
  theme(axis.text = element_text(size = 12, color = "black", face = "bold"))
```

You can see that, as the dose of either supplement increases, the teeth length also increases. 
So for both supplements, higher dose = longer teeth. 

But if you look carefully at the lines, you will notice that the two curves are not exactly parallel. 
While teeth are longer for OJ than VC in lower doses, there is no difference in dose = 2. 

And THAT is interaction. The effect of supplement depends on the dose. 
The effect of one factor (supplement) depends on the other factor (does). That's interaction. 
In the graphical sense, it means the trends are not parallel to each other. 

When you set up a multifactorial design, you will most certainly encounter interactions. 


#How to analyze a multifactorial experiment? 

You will analyze a multifactorial experiment like a randomized block experiment, 
except that you will have to consider interaction now. 

Let's fill out the table.


|  source | factor | levels |    
|  supp   | treat  |  2     |   
|  dose   | treat  |  3     |   
|  animal | OU, EU |  60    |   

We have supp and dose as our treatments. 
They have 2 and 3 levels, repectively. 
We have 60 observations of animal (each dot on the graph is an animal). 
In this case the animal is both observational units and experimental units. 

Last time we learned that we should include all sources of variation except OU. 
So that gives len ~ supp + dose. 

But how do I include the interaction between supp and dose in my linear model? 


##Setting up linear model for multifactorial design
To include an interaction in a linear model, 
you use the "factor1:factor2" syntax. 

That is len ~ supp + dose + supp:dose   

A simplified syntax is len ~ supp * dose. 
The "*" symbol means + and interaction. 
That is supp * dose = supp + dose + supp:dose 

Oftentimes people refer to the independent variables as the "main effects". 
So supp and dose are your main effects. 

When you set up your model, you type in the main effects first, then the interactions. 

WARNING: the dose is recorded as a numeric variable in the dataset. 
In a multifactorial experiment, we need to set it as factor. 
```{r}
ToothGrowth_new <- ToothGrowth %>% 
  mutate(dose.f = as.factor(dose))   
```

We can pull out the summary statistics first:
In this case we will group by both supp and does.f, the two factors. 
```{r}
ToothGrowth_new %>% 
  group_by(supp, dose.f) %>% 
  summarise(mean = mean(len),
            var = var(len),
            sd = sd(len),
            n = n()) 
```
Well, the variances are not ideal, as not all of them are within a two-fold range, but all of them are at least within an order of magnitude. 
So maybe it'll be okay. 

```{r}
model_teeth <- lm(len ~ supp + dose.f + supp:dose.f, data = ToothGrowth_new) 
```
You can also do len ~ supp * dose.f. That gives you the same thing. 

Note that you can also set it up as len ~ dose.f * supp. 
Technically there is a minute difference between the two models, 
but oftentimes it makes no difference to your conclusions at the end of the day.
We will ignore that technicality for now.


Let's check the assumptions of ANOVA first.
Normality?
```{r}
plot(model_teeth, which = 2)
```

Pretty good!

Next, equal variance? 
```{r}
plot(model_teeth, which = c(1, 3))
```

Looks rather okay to me. 
We are all clear to do ANOVA
 

##ANOVA
```{r}
anova(model_teeth)
```

1) supp has a F of 12, which is >> 1. That means the means of supp are different

2) dose.f has F of 133. This means not only are the means of dose.f different, 
but also dose.f explain a larger proportion of variation in the data than supp.

3) Lastly, the interaction supp:dose.f has a F of 5, also significantly > 1 (p = 0.02 < 0.05). 
This means the effect of supp depends on the dose, consistent with what the graph showed us. 


Now we can go ahead and find out which supp:dose combination gave us the longest teeth. 

##Tukey tests
When you do a Tukey test on a multifactorial experiment, You need to be VERY THOUGHTFUL.
Ask yourself what is the comparison that I am actually interested in? 

In this example:
Am I interested in the effect of supp at each dose level? Or,
Am I interested in the effect of dose for each supp? 

Depending on the question of interest, you will need a different Tukey test layout. 
Of course the choice of analysis all depends on the researcher and their interests. 

###test effect of supp at each dose level
Let's say we are interested in the effect of supp at each dose level. 
To specify that in the emmeans command, we will use the ~ supp | dose syntax. 
This means compare across supp at each dose level.


```{r}
estimate_per_dose <- emmeans(model_teeth,pairwise ~ supp | dose.f) 
estimate_per_dose$contrast
```

You can also pull out the cld
```{r}
cld(estimate_per_dose$emmeans, Letters = letters)
```

This tells us that at dose = 0.5 or 1, OJ gives longer teeth than VC.
However, at dose = 2, OJ and VC have about the same effect. 


###test effect of dose at each supp
Now let's say we are interested in the effect of different doses for each supp.
How should we set up the emmeans command? 

You'll use ~ dose.f | supp now. 
This means compare across dose at each supp level.


```{r}
estimate_per_supp <- emmeans(model_teeth,pairwise ~ dose.f | supp) 
estimate_per_supp$contrast
```

You can also pull up the cld
```{r}
cld(estimate_per_supp$emmeans, Letters = letters)
```

This tells us that:

1) For OJ, dose 0.5 is significantly lower than dose 1 and 2. 
Dose 2 is not significanty different from dose 1. 

2) For VC, as the dose increases, each dose is significantly different from the previous. 

Hopefully you can see that depending on how we set up the emmeans command, we get different inferences.
So before you set up a Tukey test for a multifactorial experiment,
do think about what you are actually interested in. 


#Exercise
Now you have learned how to analyze multifactorial experiment. 
You will practise it. 

We will use an actual experiment "potato yield" as example. 
In this experiment, we are looking at the yield of two varieties of potato across differet nitrogen input. 

Potato from one of the two varieties were randomly assigned one of the 5 nitrogen treatments. 
At the end of the experiment, potatos were harvested and weighed. 
(Data from UC Davis Plant Sciences course PLS205, Winter 2017) 

```{r}
Potato_Yield <- read.csv("data/Potato_Yield.csv")
head(Potato_Yield)
str(Potato_Yield)
```


##visualize the data
make the best plot you can to visualize the data.

What to put on x axis?
What to color/fill with?
Any faceting that you want to use? 

WARNING: are there any numerically recorded variables you want to change into factors?
(Hint: there is 1 variable that you need to change into factor.) 

Make your plot here:
```{r}

```





##setting up the model

Fill out the table: 

What are the sources of variation?
What factors are they? (treatment? block? observational units? experimental units?)
How many levels in each?


| source | factor | levels |   
|  ...   | ...    | ...    |   
 
What main effects should you include in the model? 
Which interactions should you include in the model? 

Set up the linear model. 
```{r}

```



##check assumptions of ANOVA 
```{r}

```

Normality?

Equal variance? 

##run ANOVA
How would you interpret the ANOVA table?
```{r}

```


##Tukey test
Say we are interested in knowing which variety performs better each nitrogen input. 
How should you set up the Tukey test? 
```{r}

```

 
Pull out the cld output and interpret the results. 
```{r}

```

 `
 






 
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
